# Tuesday, August 23, 2022 {.unnumbered}

**"Any fool can write code that a computer can understand. Good programmers write code that humans can understand."** --- Martin Fowler

## Today {.unnumbered}

In this lesson I tell you what this course is about. In do so I give you a bit of history on my journey into data science. I give details about lessons, assignments, and grading, and about how to get the most out of this course. I will talk a bit about why I present the course in this way.

-   Is Milwaukee snowier than Madison?

-   Is global warming making hurricanes stronger?

-   Are tornadoes more likely to form over smooth terrain?

-   Understand what this course is about, how it is structured, and what I expect from you

-   Getting set to work with R and RStudio

## Install R and RStudio on your computer {.unnumbered}

First get R

-   Go to <http://www.r-project.org>
-   Select the CRAN (Comprehensive R Archive Network). Scroll to a mirror site
-   Choose the appropriate file for your computer
-   Follow the instructions to install R

Then get RStudio

-   Go to on <http://rstudio.org>
-   Download RStudio Desktop
-   Install and open RStudio

Finally (not required for success in this class), learn git with R

<https://happygitwithr.com/install-git.html>

## Download course materials {.unnumbered}

-   Navigate to \<[[https://github.com/jelsner/ASS-2022\\\\](https://github.com/jelsner/ASS-2022\){.uri}]([https://github.com/jelsner/ASS-2022\\](https://github.com/jelsner/ASS-2022)%7B.uri%7D){.uri}\>
-   Click on the bright green Code button
-   Download ZIP
-   Unzip the file on your computer
-   Open the `ASS-2022.Rproj` file

## Read the syllabus {.unnumbered}

-   Open the `00-Syllabus.Rmd` file under the `Files` tab

## About RStudio {.unnumbered}

-   Written in HTML (like your Web browser)

-   Top menus

    -   File \> New File \> R Markdown
    -   Tools \> Global Options \> Appearance

-   Upper left panel is the markdown file. This is where you put your text and code

    -   Run code chunks from this panel
    -   Output from the operations can be placed in this panel or in the Console (see the gear icon above)
    -   All the text, code, and output can be rendered to an HTML file or a PDF or Word document (see the Knit button above)

-   Upper right panel shows what is in your current environment and the history of the commands you issued

    -   This is also where you can connect to github

-   Lower left panel is the Console

    -   I think of this as a sandbox where you try out small bits of code. If it works and is relevant to what you want to do you move it to the markdown file
    -   This is also where output from running code will be placed
    -   Not a place for plain text

-   Lower right panel shows your project files, the plots that get made, and all the packages associated with the project

    -   The File tab shows the files in the project. The most important one is the .Rmd.
    -   The Plot tab currently shows a blank sheet
    -   The Packages tab shows all the packages that have been downloaded from CRAN and are associated with this project

## Lab assignments {.unnumbered}

You will do all assignments inside a Rmd file.

1.  Get the assignment `Rmd` file from github and rename it to `yourLastName_yourFirstName.Rmd`
2.  Open the `Rmd` file with RStudio
3.  Replace 'Your Name' with your name in the preamble (YAML)
4.  Answer the questions by typing appropriate code between the code-chunk delimiters
5.  Select the Knit button to generate an HTML file
6.  Fix any errors
7.  Email your completed assignment `Rmd` file to [jelsner\@fsu.edu](mailto:jelsner@fsu.edu){.email}

## Getting started with R {.unnumbered}

Applied statistics is the analysis and modeling of data. Use the `c()` function to input small bits of data into R. The function combines (concatenates) items in a list together.

For example, consider a set of hypothetical annual land falling hurricane counts over a ten-year period.

2 3 0 3 1 0 0 1 2 1

You save these 10 integer values in your working directory by typing them into the console as follows. The console is the lower left window.

```{r}
counts <- c(2, 3, 0, 3, 1, 0, 0, 1, 2, 1)
counts
```

You assign the values to an object called `counts`. The assignment operator is an equal sign (`<-` or `=`). Values do not print. They are assigned to an object name. They are printed by typing the object name as we did on the second line. When printed the values are prefaced with a `[1]`. This indicates that the object is a vector and the first entry in the vector has a value of 2 (The number immediately to the right of `[1]`).

Use the arrow keys to retrieve previous commands. Each command is stored in the history file. The up-arrow key moves backwards through the history file. The left and right arrow keys move the cursor along the line.

Then you apply functions to data stored in an object.

```{r}
sum(counts)
length(counts)
sum(counts)/length(counts)
mean(counts)
```

The function `sum()` totals the number of hurricanes over all ten years, `length()` gives the number of elements in the vector. There is one element (integer value) for each year, so the function returns a value of 10.

Other functions include `sort()`, `min()`, `max()`, `range()`, `diff()`, and `cumsum()`. Try these functions on the landfall counts. What does the `range()` function do? What does the function `diff()` do?

```{r}
diff(counts)
```

The hurricane count data stored in the object `counts` is a vector. This means that R keeps track of the order that the data were entered. There is a first element, a second element, and so on. This is good for several reasons.

The vector of counts has a natural order; year 1, year 2, etc. You don't want to mix these. You would like to be able to make changes to the data item by item instead of entering the values again. Also, vectors are math objects so that math operations can be performed on them.

For example, suppose `counts` contain the annual landfall count from the first decade of a longer record. You want to keep track of counts over other decades. This is done here as follows.

```{r}
d1 <- counts
d2 <- c(0, 5, 4, 2, 3, 0, 3, 3, 2, 1)
```

Most functions operate on each element of the data vector at the same time.

```{r}
d1 + d2
```

The first year of the first decade is added from the first year of the second decade and so on.

What happens if you apply the `c()` function to these two vectors? Try it.

```{r}
c(d1, d2)
```

If you are interested in each year's count as a difference from the decade mean, you type

```{r}
d1 - mean(d1)
```

In this case a single number (the mean of the first decade) is subtracted from a vector. The result is from subtracting the number from each entry in the data vector.

This is an example of data recycling. R repeats values from one vector so that the vector lengths match. Here the mean is repeated 10 times.

## Are you completely new to R? {.unnumbered}

The {swirl} package contains functions to get you started with the basics of R. To install the package use the `install.packages()` function with the name of the package in quotes. The function downloads the package from the Comprehensive R Archive Network (CRAN). You update packages using `update.packages()` function.

To make the functions work in your current session use the `library()` function with the name of the package (without quotes). This needs to be done for every session, but only once per session.

```{r, eval=FALSE}
install.packages("swirl")
library(swirl)
```

Type:

```{r, eval=FALSE}
swirl()
```

Choose the lesson: R Programming. Work through lessons 1:8

Getting help: <https://www.r-project.org/help.html>
