# Tuesday, August 23, 2022 {-}

**"Any fool can write code that a computer can understand. Good programmers write code that humans can understand."** --- Martin Fowler

In this lesson I tell you what this course is about. In do so I give you a bit of history on my journey into data science. I give details about lessons, assignments, and grading, and about how to get the most out of this course. At the end I will talk a bit about why I present the course in this way.

* Is Milwaukee snowier than Madison?
* Is global warming making hurricanes stronger?
* Are tornadoes more likely to form over smooth terrain?

Applied statistics is the analysis and modeling of data. Use the `c()` function to input small bits of data into R. The function combines (concatenates) items in a list together. For example, consider a set of hypothetical annual land falling hurricane counts over a ten-year period.

2  3  0  3  1  0  0  1  2  1

We get save these counts in our working directory by typing them into the console as follows. The console is the lower left window.
```{r}
counts <- c(2, 3, 0, 3, 1, 0, 0, 1, 2, 1)
counts
```

We assign the values to an object called `counts`. The assignment operator is an equal sign (`<-` or `=`).  Values do not print. They are assigned to an object name. They are printed by typing the object name as we did on the second line. When printed the values are prefaced with a `[1]`. This indicates that the object is a vector and the first entry in the vector has a value of 2 (The number immediately to the right of `[1]`).

We use the arrow keys to retrieve previous commands. Each command is stored in the history file. The up-arrow key moves backwards through the history file. The left and right arrow keys move the cursor along the line.

We apply functions to data stored in objects.
```{r}
sum(counts)
length(counts)
sum(counts)/length(counts)
mean(counts)
```

The function `sum()` totals the number of hurricanes over all years, `length()` gives the number of elements in the vector. Other functions include, `sort()`, `min()`, `max()`, `range()`, `diff()`, and `cumsum()`. Try these functions on the landfall counts. What does `range()` function do?  What does `diff()` do?

The hurricane count data stored in the object `counts` is a vector. This means that R keeps track of the order that the data were entered. There is a first element, a second element, and so on. This is good for several reasons.

The vector of counts has a natural order; year 1, year 2, etc. We don't want to mix these. We would like to be able to make changes to the data item by item instead of entering the values again. Also, vectors are math objects so that math operations can be performed on them.

For example, suppose `counts` contain the annual landfall count from the first decade of a longer record.  We want to keep track of counts over other decades. This could be done by the following, example.
```{r}
d1 <- counts
d2 <- c(0, 5, 4, 2, 3, 0, 3, 3, 2, 1)
```

Most functions operate on each element of the data vector at the same time.
```{r}
d1 + d2
```

The first year of the first decade is added from the first year of the second decade and so on.

What happens if we apply the `c()` function to these two vectors?  Try it.
```{r}
c(d1, d2)
```

If we are interested in each year's count as a difference from the decade mean, we type:
```{r}
d1 - mean(d1)
```

In this case a single number (the mean of the first decade) is subtracted from a vector. The result is from subtracting the number from each entry in the data vector. This is an example of data recycling. R repeats values from one vector so that the vector lengths match. Here the mean is repeated 10 times.