# Thursday, January 12, 2023 {-}

**"Any fool can write code that a computer can understand. Good programmers write code that humans can understand."** --- Martin Fowler

## Important points from the syllabus {-}

Student Hours: TR 9:15-9:45 a.m., MW 2:15-3 p.m. The best way to get a hold of me is through email: [jelsner@fsu.edu](mailto:jelsner@fsu.edu)

This course is a survey of methods to describe, analyze, and model geographic data using R. Focus is on applications. Emphasis is on how spatial statistical methods are related through the concept of spatial autocorrelation

During each lesson I'll work through and explain the R code within an `xx-Lesson.Rmd` file. The notes in the files are comprehensive, so you can work through them on your own

Grades are determined by how well you do on the lab assignments. Lab assignments will be started in class on the following dates

| Lab | Date                  | Lessons covered |
|-----|-----------------------|-----------------|
| 1   | Thursday January 26   | 1-4             |
| 2   | Tuesday February 14   | 5-8             |
| 3   | Tuesday March 7       | 9-13            |
| 4   | Tuesday April 11      | 14-19           |
| 5   | Tuesday April 25      | 20-22           |

Lab assignments are due by 6 p.m. on that day

## Continue with setup {-}

After installing R and RStudio, install and configure Git

-   [Install & configure Git](https://happygitwithr.com/install-git.html) 

Get a GitHub account

-   [Signup for a GitHub account](https://github.com/)
-   Click on `Sign up`
-   Enter your email, Create a password, Enter a username
-   Solve the puzzle
-   Enter the code sent to your email
-   How many team members: 1, Student
-   Collaborative coding
-   Continue for free

Create a fork of the class repository on GitHub

-   [Create a fork](https://github.com/jelsner/)
-   Click on the link `jelsner/ASS-2023`
-   In the upper right click on scroll down arrow next to `Fork`
-   Select "+ Create new fork"
-   Add your name (no spaces) to the `Repository name`
-   Click on the green `Create fork`

Create a directory on your computer of your forked GitHub repository

-   Open RStudio
-   Under `File` > `New Project` select `Version Control`
-   Select `Git`
-   Repository URL: `https://github.com/[your GitHub username]/ASS-2023-[your name].git`
-   Project directory name: `ASS-2023-[your name]`
-   Create a project as a subdirectory of: e.g. `Browse to Desktop`
-   Select `Create Project`. You should now see a `Git` tab in one of the window panes (upper right)
-   Click on the `01-Lesson.Rmd`
-   Install {markdown} and {bookdown}
-   Select `Knit` to generate the lesson as HTML

At the start of each class (Sync from GitHub then Pull from RStudio)

-   Go to `https://github.com/[your GitHub username]`
-   Select the `Sync fork` button under the green `Code` button 
-   Click `Update branch`
-   Open the Folder `ASS-2023-[your name]` on your Desktop
-   Open the file `ASS-2023.Rproj`
-   Under the `Git` tab select blue arrow (pointing down) to `Pull` down changes from your GitHub repository
-   Select the `XX-Lesson.Rmd` file

Don't push!

-   You can add notes to the lesson files (`_.Rmd`) on your computer
-   You can add new files to your local directory
-   These changes stay on your computer only
-   Don't push them to your GitHub repository!

Don't panic!

If you can't get things to work with your GitHub repository and/or your RStudio does not recognize git then you can download my entire class repository all at once

-   [Don't panic](https://github.com/jelsner/ASS-2023)
-   Select the Green code button and scroll to `Download ZIP`
-   Unzip the file on your Desktop
-   From the `ASS-2023-main` folder open the `ASS-2023.Rproj` file
-   Select the `xx-lesson.Rmd` file
-   Select `Knit` to generate HTML

Getting individual files off GitHub (only if you don't have a git tab). To keep the `ASS-2023-main` folder up to date you can get the individual lesson file off GitHub

-   [Don't panic](https://github.com/jelsner/ASS-2023)
-   Select the individual lesson file (`xx-Lesson.rmd`)
-   Select the Raw tab then Save page as
-   Move that file to your folder `CCS-2023-main`

At the end of the semester you can push your notes to your repository

-   Check the `Staged` button under the `Git` tab for all files you want to push
-   Then select the `Commit` button
-   Add a note to the `Commit message` section (e.g., my updates)
-   Select the green arrow (pointing up) to `Push` changes to your GitHub
-   The _first time_ you do this you need to include your GitHub username and Personal access token

Get a personal access token

-   Go to your GitHub account
-   In the far upper-right pull down menu (next to your avatar) select `Settings` then `Developer settings`
-   In the left menu select `Personal access tokens > Tokens (classic)`
-   Follow instructions to generate a personal access toke (PAT)
-   Use your PAT as your password when pushing changes made on your computer to your GitHub account

Finally you can unfork your repository (see the instructions here <https://nickymeuleman.netlify.app/blog/unfork-github-repo>)

Help with all aspects of git, GitHub, and R programming is widespread across the web. A list of some of the sites related to spatial analysis and modeling using R are listed in the syllabus

I will spend the first several lessons teaching you how to work with R. This material might be a review. I will start by showing you how to work with small bits of data using functions from the {base} packages. The {base} packages are included in your initial installation. The functions form these packages provide the scaffolding for working with the code, but much of what you will do in this class involve functions from other packages. If this is entirely new, don't get discouraged. This class will not involve writing complex code

[Getting started with R and R-spatial](https://bakaniko.github.io/foss4g2022-getting-started-rspatial/)

## About RStudio {-}

-   Written in HTML
-   Top menus
    -   File > New File > R Markdown
    -   Pane layout (see the icon next to `Addins`)
    -   Tools > Global Options > Appearance
    
-   Upper left panel is the markdown file. This is where you put your text and code
    -   Run code chunks from this panel
    -   Output from the operations can be placed in this panel or in the Console (see the gear icon above)
    -   All the text, code, and output can be rendered to an HTML, PDF or Word (see the Knit button above)
    
-   Upper right panel shows what is in your current environment and the history of the commands you issued
    -   This is also where you can connect to github

-   Lower left panel is the Console
    -   I think of this as a sandbox where you try out small bits of code. If it works and is relevant to what you want to do you move it to the markdown file
    -   This is also where output from running code will be placed
    -   Not a place for plain text

-   Lower right panel shows your project files, the plots that get made, and all the packages associated with the project
    -   The File tab shows the files in the project
    -   The Plot tab currently shows a blank sheet
    -   The Packages tab shows all the packages that have been downloaded from CRAN and are associated with this project. Checked boxes indicate the package is loaded for this session

## Lab assignments {-}

You will do all assignments inside a `Rmd` file

1.  Get the assignment `Rmd` file from GitHub and rename it to `yourLastName_yourFirstName.Rmd`
2.  Open the `Rmd` file with RStudio
3.  Replace 'Your Name' with your name in the preamble (YAML)
4.  Answer the questions by typing appropriate code between the code-chunk delimiters
5.  Select the Knit button to generate an HTML file
6.  Fix any errors
7.  Email your completed assignment `Rmd` file (not the `html`) to [jelsner@fsu.edu](mailto:jelsner@fsu.edu)

## Data science workflow with R markdown {-}

A scientific paper is *advertisement* for a claim about the world. The *proof* is the procedure that was used to obtain the result that under girds the claim. Computer code is typically the exact procedure

Computer code is a recipe for what was done. It is the most efficient way to communicate precisely the steps involved. Communication to others and to your future self

When you use a spreadsheet to obtain a result, it's hard to explain to someone precisely what you did. Click here, then right click here, then choose menu X, etc. The words used to describe these types of procedures are not standard

If you've ever made a map using GIS you know how hard it is to make another (even similar one) with a new set of data. Running code to produce another map with new data is simple

Code is an efficient way to communicate because all the information is given as plain text without ambiguity. Being able to code is a key skill for most technical jobs

The person most likely to reproduce your work a few months later is you. This is especially true for graphs and figures. These often have a finished quality to them as a result of tweaking and adjustments to the details. This makes it hard to reproduce later. The goal is to do as much of this tweaking as possible with the code you write, rather than in a way that is invisible (retrospectively). Contrast editing an image in Adobe Illustrator

When doing data science you switch between

(1) Writing code: Code to get your data into R, code to look at tables and summary statistics, code to make graphs, code to compute statistics, code to model and plot your results

(2) Looking at output that your code produced: Your code is a set of instructions that produces the output you want: a table, a model, or a figure. It is helpful to be able to see that output

(3) Taking notes: You also write text about what you are doing, why you are doing it, and what the results mean

To be efficient you write code and comments *together* in the same file. This is where R markdown comes in (files that end with `.Rmd`). An R markdown file is a plain text document where text (such as notes or discussion) is interspersed with pieces, or chunks, of R code. When you `Knit` this file the code is executed (from the top to the bottom of the file) and the results supplement or replace the code with output

The resulting file is converted into HTML or into a PDF or Word document. Some of the text in a markdown file is for format instructions. The following symbols are used for emphasis *italics*, **bold**, and `code font`. When you create a *new* markdown document in R Studio, it contains a sample example

File > New File > R Markdown > My Rmd > Save to Desktop > Knit

Lesson notes for this class are written in text using markdown formatting as needed. Text is interspersed with code. The format for code chunks is

```{r}
z <- "hello world"# lines of code here
z
```
Three back-ticks (on a U.S. keyboard, the character under the escape key) followed by a pair of curly braces containing the name of the computer language you are using. The back-ticks-and-braces part signal that code is about to begin. You write code as needed, and then end the chunk with a new line containing three more back-ticks.

You can use the Green insert button above to save time.

In the markdown file, the lines between the first and second set of back ticks is grayed and a few small icons are noted in the upper-right corner of the grayed area. The green triangle is used to execute the code and either post the results in the console below or in the line below

When you keep notes in this way, you are able to see everything together, the code, the output it produces, and your commentary or clarification on it. Also you can turn it into a good-looking document with one click. This is how you will do everything in this course

Finally, note the *Outline* button in the upper right corner of the markdown file. You can organize and navigate through the markdown file section by section based on the pound symbol (`#`)

## Getting started with R {-}

Applied statistics is the analysis and modeling of data. Use the `c()` function to input small bits of data into R. The function combines (concatenates) items in a list together

For example, consider a set of hypothetical annual land falling hurricane counts over a ten-year period

2 3 0 3 1 0 0 1 2 1

You save these 10 integer values in your working directory by typing them into the console as follows. The console is the lower left window

```{r}
counts <- c(2, 3, 0, 3, 1, 0, 0, 1, 2, 1)
counts
```
You assign the values to an object called `counts`. The assignment operator is an equal sign (`<-` or `=`). Values do not print. They are assigned to an object name. They are printed by typing the object name as we did on the second line. When printed the values are prefaced with a `[1]`. This indicates that the object is a vector and the first entry in the vector has a value of 2 (The number immediately to the right of `[1]`).

Use the arrow keys to retrieve previous commands. Each command is stored in the history file. The up-arrow key moves backwards through the history file. The left and right arrow keys move the cursor along the line

Then you apply functions to data stored in an object as follows

```{r}
sum(counts)
length(counts)
sum(counts) / length(counts)
mean(counts)
```

The function `sum()` totals the number of hurricanes over all ten years, `length()` gives the number of elements in the vector. There is one element (integer value) for each year, so the function returns a value of 10

Other functions include `sort()`, `min()`, `max()`, `range()`, `diff()`, and `cumsum()`. Try these functions on the landfall counts. What does the `range()` function do? What does the function `diff()` do?

```{r}
diff(counts)
```

The hurricane count data stored in the object `counts` is a vector. This means that R keeps track of the order that the data were entered. There is a first element, a second element, and so on. This is good for several reasons

The vector of counts has a natural order; year 1, year 2, etc. You don't want to mix these. You would like to be able to make changes to the data item by item instead of entering the values again. Also, vectors are math objects so that math operations can be performed on them

For example, suppose `counts` contain the annual landfall count from the first decade of a longer record. You want to keep track of counts over other decades. This is done here as follows

```{r}
d1 <- counts
d2 <- c(0, 5, 4, 2, 3, 0, 3, 3, 2, 1)
```

Most functions operate on each element of the data vector at the same time

```{r}
d1 + d2
```

The first year of the first decade is added from the first year of the second decade and so on

What happens if you apply the `c()` function to these two vectors? Try it

```{r}
c(d1, d2)
```

If you are interested in each year's count as a difference from the decade mean, you type

```{r}
d1 - mean(d1)
```

In this case a single number (the mean of the first decade) is subtracted from a vector. The result is from subtracting the number from each entry in the data vector

This is an example of data recycling. R repeats values from one vector so that the vector lengths match. Here the mean is repeated 10 times

## Completely new to R? {-}

The {swirl} package contains functions to get you started with the basics of R. To install the package use the `install.packages()` function with the name of the package in quotes. The function downloads the package from the Comprehensive R Archive Network (CRAN). You update packages using `update.packages()` function

To make the functions work in your current session use the `library()` function with the name of the package (without quotes). This needs to be done for every session, but only once per session

```{r, eval=FALSE}
install.packages("swirl")
library(swirl)
```

Type:

```{r, eval=FALSE}
swirl()
```

Choose the lesson: R Programming. Work through lessons 1:8

Getting help: <https://www.r-project.org/help.html>

Ten simple rules for teaching yourself R <https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1010372#sec007>