---
title: "Lab 1"
author: "Your Name"
date: January 26, 2023
---

Due: 3 p.m.

Total Points: 75

**1** Assign to the vector object `rn` 100 random numbers from the standard normal distribution using the code below then answer the questions using {base} R code.

```{r}
rn <- rnorm(n = 100)
```

a. What is the value in the 25th element (2)
```{r}
rn[25]
```

b. What are the highest and lowest values (2)
```{r}
range(rn)
```

c. What are the mean and median values? (2)
```{r}
mean(rn)
median(rn)
```

d. How many values are less than 0? (2)
```{r}
sum(rn < 0)
```

e. How many values are between 1 and 2? (2)
```{r}
sum(rn >= 1 & rn <= 2)
```

**2** The Palmer penguins data set is available from https://raw.githubusercontent.com/allisonhorst/palmerpenguins/master/inst/extdata/penguins.csv

a. Import the csv file as a data frame with the name `penguins` using the `readr::read_csv()` function. (5)

```{r}
url <- "https://raw.githubusercontent.com/allisonhorst/palmerpenguins/master/inst/extdata/penguins.csv"
penguins <- readr::read_csv(url)
```

b. Print the number of rows and columns in the data frame. (5)
```{r}
dim(penguins)
```

c. Print the column names. (5)
```{r}
names(penguins)
```

d. How many missing values are there in the vector of bill lengths (`bill_length_mm`)? (5)
```{r}
sum(is.na(penguins$bill_length_mm))
```

e. Compute the average bill length (in mm) for all penguins in the data frame. (5)
```{r}
mean(penguins$bill_length_mm, 
     na.rm = TRUE)
```


**3** The object `us_states` from the {spData} package is a data frame from the U.S. Census Bureau. The variables include the state `GEOID` and `NAME`, the `REGION` (`South`, `West`, etc), `AREA` (in square km), and total population in 2010 (`total_pop_10`) and in 2015 (`total_pop_15`). 

a. Create a new data frame called `df` containing only the population columns. Note: you might need the `sf` package for this to work. (5) 
```{r}
library(sf)

( df <- spData::us_states |>
  dplyr::select(total_pop_10, total_pop_15) )
```

b. Create a new data frame containing only states from the South region. (5)
```{r}
( df <- spData::us_states |>
  dplyr::filter(REGION == "South") )
```


c. Create a new data frame containing only states from the West region with a 2015 population of more than 6 million people. (5)
```{r}
library(sf)

( df <- spData::us_states |>
  dplyr::filter(REGION == "West" &
                total_pop_15 > 6e6) )

```


d. What was the total population of the Midwest region in 2010 and 2015? (5)

```{r}
spData::us_states |>
  dplyr::filter(REGION == "Midwest") |>
  dplyr::summarize(totPop2010 = sum(total_pop_10),
                   totPop2015 = sum(total_pop_15))
```

e. How many states are in each region and what is the _average_ statewide 2015 population in each region? Hint: Use the `dplyr::group_by()` function. (5)

```{r}
spData::us_states |>
  sf::st_drop_geometry() |>
  dplyr::group_by(REGION) |>
  dplyr::summarize(nS = dplyr::n(),
                   Avg = mean(total_pop_15))
```

**4** Use the file `Tornadoes.csv` in the directory `data` and create a time series plot of the annual number of tornadoes occurring in Florida since 1950. Use a point and a line layer (15).
```{r}
library(ggplot2)

readr::read_csv(file = here::here("data", "Tornadoes.csv")) |>
  dplyr::filter(st == "FL") |>
  dplyr::group_by(yr) |>
  dplyr::summarize(nT = dplyr::n()) |>
  ggplot(mapping = aes(x = yr, y = nT)) +
  geom_line() +
  geom_point()
```

