# Tuesday, January 17, 2023 {-}

**"If you give your code to a competent programmer who's never seen it and they ask any questions, you've failed. Every question is a defect."** --- Allen Holub

## Class notes {-}

You should now have a set of class notes as lesson files in your own GitHub repository and on your laptop. Note that these files are based on last semester's class. I will update them as we proceed through the semester

So at the start of each class (Sync from GitHub then Pull from RStudio)

-   Go to `https://github.com/[your GitHub username]`
-   Select the `Sync fork` button under the green `Code` button 
-   Click `Update branch`

-   Open the Folder `ASS-2023-[your name]` on your Desktop
-   Open the file `ASS-2023.Rproj`
-   Under the `Git` tab select blue arrow (pointing down) to `Pull` down changes from your GitHub repository
-   Select the `XX-Lesson.Rmd` file

Note: If you see a yellow caution sign when you open the lesson file, download the required packages

Don't push and rename your files

-   You can add notes to the lesson files (`_.Rmd`) on your computer
-   You can add new files to your local directory
-   These changes stay on your computer only
-   Don't push them to your GitHub repository!

At the end of the semester you can push your notes to your repository

-   Check the `Staged` button under the `Git` tab for all files you want to push
-   Then select the `Commit` button
-   Add a note to the `Commit message` section (e.g., my updates)
-   Select the green arrow (pointing up) to `Push` changes to your GitHub
-   The _first time_ you do this you need to include your GitHub username and Personal access token

Get a personal access token

-   Go to your GitHub account
-   In the far upper-right pull down menu (next to your avatar) select `Settings` then `Developer settings`
-   In the left menu select `Personal access tokens > Tokens (classic)`
-   Follow instructions to generate a personal access toke (PAT)
-   Use your PAT as your password when pushing changes made on your computer to your GitHub account

Finally you can unfork your repository (see the instructions here <https://nickymeuleman.netlify.app/blog/unfork-github-repo>)

New resource for applied spatial statistics <https://geocompx.org/>

## Using R {-}

Consider a set of integers representing the number of hurricanes making landfall in the United States from the last ten years. You assign the values to an object and print the contents of the object by wrapping the line of code in parentheses

```{r}
( d <- c(2, 3, 0, 3, 1, 0, 0, 1, 2, 1) )
```

Suppose you are interested in the inter annual variability in the set of landfall counts. The variance is computed as

$$
\hbox{var}(x) = \frac{(x_1 - \bar x)^2 + (x_2 - \bar x)^2 + \cdots + (x_n - \bar x)^2}{n-1} = \frac{1}{n-1}\sum_{i=1}^n (x_i - \bar x)^2
$$

Note: the markdown file recognizes math written in LaTeX. The formula for the variance is written using LaTeX notation gets rendered as a math equation (like you see in a textbook) See <https://www.overleaf.com/learn/latex/Mathematical_expressions>

Although the `var()` function computes this, here you do this using simple functions. The key is to find the squared differences and then sum these values

```{r}
x <- d
xbar <- mean(x)
x - xbar
(x - xbar)^2
sum((x - xbar)^2)
n <- length(x)
n
sum((x - xbar)^2)/(n - 1)
var(x)
```

Elements in a vector object must all have the same type. This type can be numeric, as in counts, character strings, as in

```{r}
simpsons <- c('Homer', 'Marge', 'Bart', 'Lisa', 'Maggie')
simpsons
```

Character strings are made with matching quotes, either double, `"`, or single, `'`. If you mix types the values will be coerced into a common type, which is usually a character string. Arithmetic operations do not work on character strings

Returning to the land falling hurricane counts. Suppose the National Hurricane Center (NHC) reanalyzes a storm, and that the 6th year of the decade is a 1 rather than a 0 for the number of landfalls. In this case you change the sixth element to have the value 1

```{r}
d[6] <- 1
```

You assign to the 6th year of the decade a value of one. The square brackets `[]` are used to reference elements of the data vector

Key: Parentheses `()` are used by functions and square brackets `[]` are used by data objects

```{r}
d
d[2]
d[-4]
d[c(1, 3, 5, 7, 9)]
```

The first line prints all the elements of the vector `d`. The second prints only the 2nd value of the vector. The third prints all but the 4th value. The fourth prints the values with odd element numbers

To create structured data, for example the integers 1 through 99 you can use the `:` operator

```{r, eval=FALSE}
1:99
rev(1:99)
99:1
```

The `seq()` function is more general. You specify the sequence interval with the `by =` or `length =` arguments

```{r}
seq(from = 1, to = 9, by = 2)
seq(from = 1, to = 10, by = 2)
seq(from = 1, to = 9, length = 5)
```

The `rep()` function is used to create repetitive sequences. The first argument is a value or vector that you want repeated and the second argument is the number of times you want it repeated

```{r}
rep(1, times = 10)
rep(simpsons, times = 2)
```

In the second example the vector `simpsons` containing the Simpson characters is repeated twice

To repeat each element of the vector use the `each =` argument

```{r}
rep(simpsons, each = 2)
```

More complicated patterns can be repeated by specifying pairs of equal length vectors. In this case, each element of the first vector is repeated the corresponding number of times given by the element in the second vector

```{r}
rep(c("long", "short"),  times = c(2, 3))
```

To find the maximum number of landfalls during the first decade you type

```{r}
max(d)
```

What years had the maximum?

```{r}
d == 3
```

Notice the double equals signs (`==`). This is a logical operator that tests each value in `d` to see if it is equal to 3. The 2nd and 4th values are equal to 3 so `TRUE`s are returned

Think of this as asking R a question. Is the value equal to 3? R answers all at once with a vector of `TRUE`'s and `FALSE`'s

What years had more than 2 hurricanes?

```{r}
d > 2
```

Now the question is how do you get the vector element corresponding to the `TRUE` values? That is, which years have 3 landfalls?

```{r}
which(d == 3)
```

The function `which.max()` can be used to get the first maximum

```{r}
which.max(d)
```

You might be interested in the total number of landfalls in the decade or the number of years in a decade without a landfall

```{r}
sum(d)
sum(d == 0)
```

With a logical vector the `TRUE` elements are treated as having a value of 1 and the `FALSE` elements are treated as having a value of 0

The `rm()` function removes an object from your environment

```{r}
rm(d)
```

## Data frames {-}

Spatial data frames are used throughout this course. A spatial data frame is a data frame plus information on where the observations were taken

A data frame stores data in a tabular format like a spreadsheet. It is a list of vectors each with the same length. It has column names (and sometimes row names)

For example, you create a data frame object `df` containing three vectors `n`, `s`, `b` each with three elements using the `data.frame()` function

```{r}
n <- c(2, 3, 5) 
s <- c("aa", "bb", "cc") 
b <- c(TRUE, FALSE, TRUE) 

df <- data.frame(n, s, b)
```

To see check on the type of object you use the `class()` function with the name of the object inside the parentheses

```{r}
class(df)
```

The object `df` is of class `data.frame`. Note that the object name shows up in our *Environment* under *Data* and it includes a little blue arrow indicating that you can view it by clicking on the row

The data frame shows up as a table (like a spreadsheet) in the `View()` mode (see the command in the console below). This is not advised for large data frames!

The top line of the table is called the header. Each line below the header contains a row of data, which begins with the name (or number) of the row followed by the data values

Each data element is in a cell. To retrieve a data value from a cell, you enter its row and column coordinates in that order in the single square bracket `[]` operator and separated by a column

df[row, column]

Here is the cell value from the first row, second column of `df`

```{r}
df[1, 2]
```

You print the column names (located in the top row in the `View()` mode) with the `names()` function

```{r}
names(df)
```

The list of names is a vector of length three containing the elements `n`, `s`, and `b` in that order

You access individual columns of a data frame as vectors by appending the dollar sign (`$`) to the object name. For example, to print the values of the column labeled `s` type

```{r}
df$s
```

Many packages include example data frames. The data frame called `mtcars`, for instance, contains information extracted from the 1974 Motor Trend US magazine, and comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973--74 models)

```{r}
class(mtcars)
names(mtcars)
```

The number of data rows and data columns in the data frame are printed using the `nrow()` and `ncol()` functions

```{r}
nrow(mtcars)
ncol(mtcars)
```

Further details of built-in data frames like `mtcars` is available in the documentation accessed with the `help()` (or `?`) function

```{r, eval=FALSE}
help(mtcars)
```

If you type the name of the data frame in the console all the data are printed

```{r}
mtcars
```

Instead, you use the function `head()`, which prints the first six rows, or `str()`, which lists all the columns by data type

```{r}
head(mtcars)
str(mtcars)
```

Consider the data frame `studentdata` from the {LearnBayes} package. To access this data frame, first install the package with the `install.packages()` function

You put the name of the package {LearnBayes} in quotes (single or double). Then to make the functions from the package available to your current session use the `library()` function with the name of the package (unquoted) inside the parentheses

```{r}
if(!require(LearnBayes)) install.packages(pkgs = "LearnBayes", repos = "http://cran.us.r-project.org")

library(LearnBayes)
```

Note: The argument `repos =` in the `install.packages()` function directs where the package is obtained on CRAN (comprehensive R archive network). The CRAN repository is set automatically when using RStudio and you can install packages by clicking on *Packages* > *Install* in the lower-right panel

You need to specify the repository. When you render the markdown file (using the `Knit` button)  you don't want to install packages that already exist on your computer so you add the conditional `if()` function that says "only install the package IF it is not (`!`) available"

Make a copy of the data frame by assigning it to an object with the name `df` and print the first six rows using the `head()` function

```{r}
df <- studentdata
head(df)
```

Data frames are like spreadsheets with rows and columns. Each row is an observation (here each row is a student in an intro stats class at Bowling Green State University) and each column is a variable. Here the variables are answers to questions like what is your height, choose a number between 1 and 10, what time did you go to bed last night, etc

The names of the columns are printed using the `names()` function

```{r}
names(df)
```

All columns are of the same length, but not all students answered all questions so some of the data frame cells contain the missing-value indicator `NA` (not available)

Values in a data frame stored in rows and columns and are accessed with bracket notation [row, column] where row is the row number and column is the column number like a matrix

For example here you specify the value in the 10th row and 2nd column (`Height` column) of the `df` data frame

```{r}
df[10, 2]
```

By specifying only the row index and leaving the column index blank you get all values in that row, which corresponds to all the responses given by the 10th student

```{r}
df[10, ]
```

Drink preference was one of the questions. Responses across all students are available in the column labeled `Drink` as a vector of character values. You list the various drink preferences by typing

```{r}
df$Drink
```

Some students left that response blank and therefore the response is coded with the missing-value indicator

The variable type depends on the question asked. For example, answers given to the question of student height result in a numeric variable, answers given to the question about drink preference result in a character (or factor) variable

For integer, character, and factor variables you summarize the set of responses with the `table()` function

```{r}
table(df$Drink)
```

There are 113 students who prefer milk, 178 prefer soda, and 355 prefer water

Here you use the`plot()` method to make a draft plot of this table

```{r}
plot(x = df$Drink)
```

Notice that the sum of the responses is `r sum(table(df$Drink))`, which is less than the total number of students (`r nrow(df)`).

Students who left that question blank are ignored in the `table()` function. To include the missing values you add the argument `useNA = "ifany"` to the `table()` function.

```{r}
table(df$Drink, 
      useNA = "ifany")
```

Note: When you want code executed directly within the text you separate the code using single back ticks. This is useful when you write reports that need periodic updates when new data becomes available. Instead if you hard code the values in the text then you need to search the document for these values during each update of the report

Suppose you are interested in examining how long students reported sleeping during the night. This was not asked directly. You compute it from the `ToSleep` and `WakeUp` times columns. You assign the result of the difference to a column we call `SleepHrs`

```{r}
df$SleepHrs <- df$WakeUp - df$ToSleep
head(df)
```

Now you have a new numeric variable in the data frame called `SleepHrs`

You can't table numeric variables, but the `summary()` method prints a set of summary statistics for the set of values

```{r}
summary(df$SleepHrs)
```

The average number of hours slept is 7.4 with a maximum of 12.5 and a minimum of 2.5. There are four students that did not answer either when they went to sleep or when they woke up questions

You use the `hist()` function to plot a histogram of sleep hours. The function takes as input a vector of values

```{r}
hist(df$SleepHrs)
```

The function divides the number of sleep hours into one-hour bins and counts the number of students whose reported sleep hours falls into each bin. For example based on when they said they went to sleep and when the said they woke up, about 100 students slept between five and six hours the night before the survey

Is the amount of sleep for a student related to when they go to bed? If you place numeric variables on the x and y axes then you get a scatter plot

```{r}
plot(x = df$ToSleep,
     y = df$SleepHrs)
```

The `ToSleep` variable is centered on midnight so that -2 means a student went to sleep at 10p

You describe the decreasing relationship with a line through the points. The least-squares line is fit using the `lm()` function and the line is drawn on the existing plot with the `abline()` function applied to the linear regression object `model`.

```{r}
model <- lm(SleepHrs ~ ToSleep, 
            data = df)

plot(x = df$ToSleep,
     y = df$SleepHrs)
abline(model)
```