# Tuesday, January 17, 2023 {.unnumbered}

**"The trouble with programmers is that you can never tell what a programmer is doing until it's too late."** --- Seymour Cray

## Data science workflow with R markdown {.unnumbered}

A scientific paper is *advertisement* for a claim about the world. The *proof* is the procedure that was used to obtain the result that under girds the claim. Computer code is typically the exact procedure.

Computer code is a recipe for what was done. It is the most efficient way to communicate precisely the steps involved. Communication to others and to your future self.

When you use a spreadsheet to obtain a result, it's hard to explain to someone precisely what you did. Click here, then right click here, then choose menu X, etc. The words used to describe these types of procedures are not standard.

If you've ever made a map using GIS you know how hard it is to make another (even similar one) with a new set of data. Running code to produce another map with new data is simple.

Code is an efficient way to communicate because all the information is given as plain text without ambiguity. Being able to code is a key skill for most technical jobs.

The person most likely to reproduce your work a few months later is you. This is especially true for graphs and figures. These often have a finished quality to them as a result of tweaking and adjustments to the details. This makes it hard to reproduce later. The goal is to do as much of this tweaking as possible with the code you write, rather than in a way that is invisible (retrospectively). Contrast editing an image in Adobe Illustrator.

When doing data science you switch between:

(1) Writing code: Code to get your data into R, code to look at tables and summary statistics, code to make graphs, code to compute statistics, code to model and plot your results.

(2) Looking at output that your code produced: Your code is a set of instructions that produces the output you want: a table, a model, or a figure. It is helpful to be able to see that output.

(3) Taking notes: You also write text about what you are doing, why you are doing it, and what the results mean.

To be efficient you write code and comments *together* in the same file. This is where R markdown comes in (files that end with `.Rmd`). An R markdown file is a plain text document where text (such as notes or discussion) is interspersed with pieces, or chunks, of R code. When you `Knit` this file the code is executed (from the top to the bottom of the file) and the results supplement or replace the code with output.

The resulting file is converted into HTML or into a PDF or Word document. Some of the text in a markdown file is for format instructions. The following symbols are used for emphasis *italics*, **bold**, and `code font`. When you create a *new* markdown document in R Studio, it contains a sample example.

File \> New File \> R Markdown \> My Rmd \> Save to Desktop \> Knit

Lesson notes for this class are written in text using markdown formatting as needed. Text is interspersed with code. The format for code chunks is

```{r}
z <- "hello world"# lines of code here
z
```

Three back-ticks (on a U.S. keyboard, the character under the escape key) followed by a pair of curly braces containing the name of the computer language you are using. The back-ticks-and-braces part signal that code is about to begin. You write code as needed, and then end the chunk with a new line containing three more back-ticks.

You can use the Green insert button above to save time.

In the markdown file, the lines between the first and second set of back ticks is grayed and a few small icons are noted in the upper-right corner of the grayed area. The green triangle is used to execute the code and either post the results in the console below or in the line below.

When you keep notes in this way, you are able to see everything together, the code, the output it produces, and your commentary or clarification on it. Also you can turn it into a good-looking document with one click. This is how you will do everything in this course.

Finally, note the *Outline* button in the upper right corner of the markdown file. You can organize and navigate through the markdown file section by section based on the pound symbol (`#`).

## An introduction to using R {.unnumbered}

Applied spatial statistics is the analysis and modeling of data that was collected across space. To begin you need to know about data objects.

The `c()` function is used to create a simple data object (vector object). The function combines (concatenates) individual values into a vector. The length of the vector is the number of data values.

Consider a set of annual land falling hurricane counts over a ten-year period. In the first year there were two hurricanes, the next year there were three, and so on.

2 3 0 3 1 0 0 1 2 1

You save these ten values by assigning them to an object that you call `counts`. The assignment operator is a left-pointing arrow or equal sign (`<-` or `=`). The left-pointing arrow is the less than symbol followed by the hyphen.

```{r}
counts <- c(2, 3, 0, 3, 1, 0, 0, 1, 2, 1)
```

By clicking on the *Environment* tab in the upper-right panel you see that the object `counts` with numerical values (`num`) 2 3, etc below word *Values*. The elements of the vector object are indexed between 1 and 10 (`1:10`).

You print the values to the console by typing the name of the data object.

```{r}
counts
```

When printed the values are prefaced with a `[1]`. This indicates that the object is a vector and the first element in the vector has a value of 2 (The number immediately to the right of `[1]`).

Note: You can assign and print by wrapping the entire line of code in parentheses.

```{r}
( counts <- c(2, 3, 0, 3, 1, 0, 0, 1, 2, 1) )
```

You can use the arrow keys on your keyboard to retrieve previous commands. Each command is stored in the history file (click on the *History* tab in the upper-right panel). The up-arrow key moves backwards through the history file. The left and right arrow keys move the cursor along the line.

You apply functions to data objects. A function has a name and parentheses. Inside the parentheses are the function arguments. Many functions have only a single argument, the data object.

```{r}
sum(counts)
length(counts)
sum(counts)/length(counts)
mean(counts)
```

The function `sum()` totals the hurricane counts over all years, `length()` returns the number of elements in the vector. Other functions include `sort()`, `min()`, `max()`, `range()`, `diff()`, and `cumsum()`.

The object `counts` that you created is a vector in the sense that the elements are ordered. There is a first element, a second element, and so on. This is good for several reasons.

The hurricane counts have a chronological order: year 1, year 2, etc and you want that ordered reflected in the data object. Also, you would like to be able to make changes to the data values by element. Also, vectors are math objects so that math operations can be performed on them in a natural way.

For example, math tells us that a scalar multiplied by a vector is a vector where each element of the product has been multiplied by the scalar. The asterisk `*` is used for multiplication.

```{r}
10 * counts
```

Further, suppose `counts` contain the annual landfall count from the first decade of a longer record. You want to keep track of counts over other decades.

```{r}
d1 <- counts
d2 <- c(0, 5, 4, 2, 3, 0, 3, 3, 2, 1)
```

Most functions operate on each element of the data vector all at once.

```{r}
d1 + d2
```

The first year of the first decade is added to the first year of the second decade and so on.

What happens if you apply the `c()` function to these two vectors?

```{r}
c(d1, d2)
```

You get a vector with elements from both `d1` and `d2` in the order of first the first decade counts and then the second decade counts.

If you are interested in each year's count as a difference from the average number over the decade you type

```{r}
d1 - mean(d1)
```

In this case a single number (the average of the first decade) is subtracted from each element of the vector.

Suppose you are interested in the inter annual variability in the set of landfall counts. The variance is computed as

$$
\hbox{var}(x) = \frac{(x_1 - \bar x)^2 + (x_2 - \bar x)^2 + \cdots + (x_n - \bar x)^2}{n-1} = \frac{1}{n-1}\sum_{i=1}^n (x_i - \bar x)^2
$$

Note: the markdown file recognizes math written in LaTeX. The formula for the variance is written using LaTeX notation gets rendered as a math equation as in a textbook. See <https://www.overleaf.com/learn/latex/Mathematical_expressions>.

Although the `var()` function computes this, here you see how to do this using simple functions. The key is to find the squared differences and then sum.

```{r}
x <- d1
xbar <- mean(x)
x - xbar
(x - xbar)^2
sum((x - xbar)^2)
n <- length(x)
n
sum((x - xbar)^2)/(n - 1)
var(x)
```

Elements in a vector object must all have the same type. This type can be numeric, as in counts, character strings, as in

```{r}
simpsons <- c('Homer', 'Marge', 'Bart', 'Lisa', 'Maggie')
simpsons
```

Character strings are made with matching quotes, either double, `"`, or single, `'`. If you mix types the values will be coerced into a common type, which is usually a character string. Arithmetic operations do not work on character strings.

Returning to the land falling hurricane counts. Now suppose the National Hurricane Center (NHC) reanalyzes a storm, and that the 6th year of the 2nd decade is a 1 rather than a 0 for the number of landfalls. In this case you change the sixth element to have the value 1.

```{r}
d2[6] <- 1
```

You assign to the 6th year of the decade a value of one. The square brackets `[]` are used to reference elements of the data vector.

It is important to keep this straight: Parentheses `()` are used by functions and square brackets `[]` are used by data objects.

```{r}
d2
d2[2]
d2[-4]
d2[c(1, 3, 5, 7, 9)]
```

The first line prints all the elements of the vector `df2`. The second prints only the 2nd value of the vector. The third prints all but the 4th value. The fourth prints the values with odd element numbers.

To create structured data, for example the integers 1 through 99 you can use the `:` operator.

```{r, eval=FALSE}
1:99
rev(1:99)
99:1
```

The `seq()` function is more general. You specify the sequence interval with the `by =` or `length =` arguments.

```{r}
seq(from = 1, to = 9, by = 2)
seq(from = 1, to = 10, by = 2)
seq(from = 1, to = 9, length = 5)
```

The `rep()` function is used to create repetitive sequences. The first argument is a value or vector that you want repeated and the second argument is the number of times you want it repeated.

```{r}
rep(1, times = 10)
rep(simpsons, times = 2)
```

In the second example the vector `simpsons` containing the Simpson characters is repeated twice.

To repeat each element of the vector use the `each =` argument.

```{r}
rep(simpsons, each = 2)
```

More complicated patterns can be repeated by specifying pairs of equal length vectors. In this case, each element of the first vector is repeated the corresponding number of times given by the element in the second vector.

```{r}
rep(c("long", "short"),  times = c(2, 3))
```

To find the maximum number of landfalls during the first decade you type

```{r}
max(d1)
```

What years had the maximum?

```{r}
d1 == 3
```

Notice the double equals signs (`==`). This is a logical operator that tests each value in `d1` to see if it is equal to 3. The 2nd and 4th values are equal to 3 so `TRUE`s are returned.

Think of this as asking R a question. Is the value equal to 3? R answers all at once with a vector of `TRUE`'s and `FALSE`'s.

What years had more than 2 hurricanes?

```{r}
d1 > 2
```

Now the question is how do you get the vector element corresponding to the `TRUE` values? That is, which years have 3 landfalls?

```{r}
which(d1 == 3)
```

The function `which.max()` can be used to get the first maximum.

```{r}
which.max(d1)
```

You might also want to know the total number of landfalls in each decade and the number of years in a decade without a landfall. Or how about the ratio of the mean number of landfalls over the two decades.

```{r}
sum(d1)
sum(d2)
sum(d1 == 0)
sum(d2 == 0)
mean(d2)/mean(d1)
```

So there are 85% more landfalls during the second decade. Is this difference statistically significant?

To remove an object from your environment use the `rm()` function.

```{r}
rm(d1, d2)
```

## Data frames {.unnumbered}

Spatial data frames will be used throughout this course. A spatial data frame is a data frame plus information about the spatial geometry. Let's start with data frames.

A data frame stores data in a tabular format like a spreadsheet. It is a list of vectors each with the same length. It has column names (and sometimes row names).

For example, you create a data frame object `df` containing three vectors `n`, `s`, `b` each with three elements using the `data.frame()` function.

```{r}
n <- c(2, 3, 5) 
s <- c("aa", "bb", "cc") 
b <- c(TRUE, FALSE, TRUE) 

df <- data.frame(n, s, b)
```

To see that the object is indeed a data frame you use the `class()` function with the name of the object inside the parentheses.

```{r}
class(df)
```

The object `df` is of class `data.frame`. Note that the object name shows up in our *Environment* under *Data* and it includes a little blue arrow indicating that you can view it by clicking on the row.

The data frame shows up as a table (like a spreadsheet) in the `View()` mode (see the command in the console below). Caution: This is not advised for large data frames.

The top line of the table is called the header. Each line below the header contains a row of data, which begins with the name (or number) of the row followed by the data values.

Each data element is in a cell. To retrieve a data value from a cell, you enter its row and column coordinates in that order in the single square bracket `[]` operator and separated by a column.

df[row, column]

Here is the cell value from the first row, second column of `df`.

```{r}
df[1, 2]
```

You can print the column names (located in the top row in the `View()` mode) with the `names()` function.

```{r}
names(df)
```

The list of names is a vector of length three containing the elements "n", "s", and "b" in that order.

You access individual columns of a data frame as vectors by appending the dollar sign (`$`) to the object name. For example, to print the values of the column labeled `s` type

```{r}
df$s
```

Many of the packages you use this semester include example data frames. The data frame called `mtcars`, for instance, contains information extracted from the 1974 Motor Trend US magazine, and comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973--74 models).

```{r}
class(mtcars)

names(mtcars)
```

The number of data rows and data columns in the data frame are printed using the `nrow()` and `ncol()` functions.

```{r}
nrow(mtcars)
ncol(mtcars)
```

Further details of built-in data frames like `mtcars` is available in the documentation accessed with the `help()` (or `?`) function.

```{r, eval=FALSE}
help(mtcars)
```

If you type the name of the data frame in the console all the data are printed.

```{r}
mtcars
```

Instead, you can use the function `head()`, which prints the first six rows, or `str()`, which lists all the columns by data type.

```{r}
head(mtcars)

str(mtcars)
```
