# Tuesday, January 17, 2023 {-}

**"If you give your code to a competent programmer who's never seen it and they ask any questions, you've failed. Every question is a defect."** --- Allen Holub

## Class notes {-}

You should now have a set of class notes as lesson files in your own GitHub repository and on your laptop. Note that these files are based on last semester's class. I will update them as we proceed through the semester

So at the start of each class (Sync from GitHub then Pull from RStudio)

-   Go to `https://github.com/[your GitHub username]`
-   Select the `Sync fork` button under the green `Code` button 
-   Click `Update branch`
-   Open the Folder `ASS-2023-[your name]` on your Desktop
-   Open the file `ASS-2023.Rproj`
-   Under the `Git` tab select blue arrow (pointing down) to `Pull` down changes from your GitHub repository
-   Select the `XX-Lesson.Rmd` file

Note: If you see a yellow caution sign when you open the lesson file, download the required packages

Don't push!

-   You can add notes to the lesson files (`_.Rmd`) on your computer
-   You can add new files to your local directory
-   These changes stay on your computer only
-   Don't push them to your GitHub repository!

At the end of the semester you can push your notes to your repository

-   Check the `Staged` button under the `Git` tab for all files you want to push
-   Then select the `Commit` button
-   Add a note to the `Commit message` section (e.g., my updates)
-   Select the green arrow (pointing up) to `Push` changes to your GitHub
-   The _first time_ you do this you need to include your GitHub username and Personal access token

Get a personal access token

-   Go to your GitHub account
-   In the far upper-right pull down menu (next to your avatar) select `Settings` then `Developer settings`
-   In the left menu select `Personal access tokens > Tokens (classic)`
-   Follow instructions to generate a personal access toke (PAT)
-   Use your PAT as your password when pushing changes made on your computer to your GitHub account

Finally you can unfork your repository (see the instructions here <https://nickymeuleman.netlify.app/blog/unfork-github-repo>)

## Using R {-}

Consider a set of integers representing the number of hurricanes making landfall in the United States from the last ten years. You assign the values to an object and print the contents of the object by wrapping the line of code in parentheses

```{r}
( d <- c(2, 3, 0, 3, 1, 0, 0, 1, 2, 1) )
```

Suppose you are interested in the inter annual variability in the set of landfall counts. The variance is computed as

$$
\hbox{var}(x) = \frac{(x_1 - \bar x)^2 + (x_2 - \bar x)^2 + \cdots + (x_n - \bar x)^2}{n-1} = \frac{1}{n-1}\sum_{i=1}^n (x_i - \bar x)^2
$$

Note: the markdown file recognizes math written in LaTeX. The formula for the variance is written using LaTeX notation gets rendered as a math equation (like you see in a textbook) See <https://www.overleaf.com/learn/latex/Mathematical_expressions>

Although the `var()` function computes this, here you do this using simple functions. The key is to find the squared differences and then sum these values

```{r}
x <- d
xbar <- mean(x)
x - xbar
(x - xbar)^2
sum((x - xbar)^2)
n <- length(x)
n
sum((x - xbar)^2)/(n - 1)
var(x)
```

Elements in a vector object must all have the same type. This type can be numeric, as in counts, character strings, as in

```{r}
simpsons <- c('Homer', 'Marge', 'Bart', 'Lisa', 'Maggie')
simpsons
```

Character strings are made with matching quotes, either double, `"`, or single, `'`. If you mix types the values will be coerced into a common type, which is usually a character string. Arithmetic operations do not work on character strings

Returning to the land falling hurricane counts. Suppose the National Hurricane Center (NHC) reanalyzes a storm, and that the 6th year of the decade is a 1 rather than a 0 for the number of landfalls. In this case you change the sixth element to have the value 1

```{r}
d[6] <- 1
```

You assign to the 6th year of the decade a value of one. The square brackets `[]` are used to reference elements of the data vector

Key: Parentheses `()` are used by functions and square brackets `[]` are used by data objects

```{r}
d
d[2]
d[-4]
d[c(1, 3, 5, 7, 9)]
```

The first line prints all the elements of the vector `d`. The second prints only the 2nd value of the vector. The third prints all but the 4th value. The fourth prints the values with odd element numbers

To create structured data, for example the integers 1 through 99 you can use the `:` operator

```{r, eval=FALSE}
1:99
rev(1:99)
99:1
```

The `seq()` function is more general. You specify the sequence interval with the `by =` or `length =` arguments

```{r}
seq(from = 1, to = 9, by = 2)
seq(from = 1, to = 10, by = 2)
seq(from = 1, to = 9, length = 5)
```

The `rep()` function is used to create repetitive sequences. The first argument is a value or vector that you want repeated and the second argument is the number of times you want it repeated

```{r}
rep(1, times = 10)
rep(simpsons, times = 2)
```

In the second example the vector `simpsons` containing the Simpson characters is repeated twice

To repeat each element of the vector use the `each =` argument

```{r}
rep(simpsons, each = 2)
```

More complicated patterns can be repeated by specifying pairs of equal length vectors. In this case, each element of the first vector is repeated the corresponding number of times given by the element in the second vector

```{r}
rep(c("long", "short"),  times = c(2, 3))
```

To find the maximum number of landfalls during the first decade you type

```{r}
max(d)
```

What years had the maximum?

```{r}
d == 3
```

Notice the double equals signs (`==`). This is a logical operator that tests each value in `d` to see if it is equal to 3. The 2nd and 4th values are equal to 3 so `TRUE`s are returned

Think of this as asking R a question. Is the value equal to 3? R answers all at once with a vector of `TRUE`'s and `FALSE`'s

What years had more than 2 hurricanes?

```{r}
d > 2
```

Now the question is how do you get the vector element corresponding to the `TRUE` values? That is, which years have 3 landfalls?

```{r}
which(d == 3)
```

The function `which.max()` can be used to get the first maximum

```{r}
which.max(d)
```

You might be interested in the total number of landfalls in the decade or the number of years in a decade without a landfall

```{r}
sum(d)
sum(d == 0)
```

With a logical vector the `TRUE` elements are treated as having a value of 1 and the `FALSE` elements are treated as having a value of 0

The `rm()` function removes an object from your environment

```{r}
rm(d)
```

## Data frames {-}

Spatial data frames are used throughout this course. A spatial data frame is a data frame plus information on where the observations were taken

A data frame stores data in a tabular format like a spreadsheet. It is a list of vectors each with the same length. It has column names (and sometimes row names)

For example, you create a data frame object `df` containing three vectors `n`, `s`, `b` each with three elements using the `data.frame()` function

```{r}
n <- c(2, 3, 5) 
s <- c("aa", "bb", "cc") 
b <- c(TRUE, FALSE, TRUE) 

df <- data.frame(n, s, b)
```

To see check on the type of object you use the `class()` function with the name of the object inside the parentheses

```{r}
class(df)
```

The object `df` is of class `data.frame`. Note that the object name shows up in our *Environment* under *Data* and it includes a little blue arrow indicating that you can view it by clicking on the row

The data frame shows up as a table (like a spreadsheet) in the `View()` mode (see the command in the console below). This is not advised for large data frames!

The top line of the table is called the header. Each line below the header contains a row of data, which begins with the name (or number) of the row followed by the data values

Each data element is in a cell. To retrieve a data value from a cell, you enter its row and column coordinates in that order in the single square bracket `[]` operator and separated by a column

df[row, column]

Here is the cell value from the first row, second column of `df`

```{r}
df[1, 2]
```

You print the column names (located in the top row in the `View()` mode) with the `names()` function

```{r}
names(df)
```

The list of names is a vector of length three containing the elements `n`, `s`, and `b` in that order

You access individual columns of a data frame as vectors by appending the dollar sign (`$`) to the object name. For example, to print the values of the column labeled `s` type

```{r}
df$s
```

Many packages include example data frames. The data frame called `mtcars`, for instance, contains information extracted from the 1974 Motor Trend US magazine, and comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973--74 models)

```{r}
class(mtcars)
names(mtcars)
```

The number of data rows and data columns in the data frame are printed using the `nrow()` and `ncol()` functions

```{r}
nrow(mtcars)
ncol(mtcars)
```

Further details of built-in data frames like `mtcars` is available in the documentation accessed with the `help()` (or `?`) function

```{r, eval=FALSE}
help(mtcars)
```

If you type the name of the data frame in the console all the data are printed

```{r}
mtcars
```

Instead, you use the function `head()`, which prints the first six rows, or `str()`, which lists all the columns by data type

```{r}
head(mtcars)
str(mtcars)
```

Consider the data frame `studentdata` from the {LearnBayes} package. To access this data frame, first install the package with the `install.packages()` function

You put the name of the package {LearnBayes} in quotes (single or double). Then to make the functions from the package available to your current session use the `library()` function with the name of the package (unquoted) inside the parentheses

```{r}
if(!require(LearnBayes)) install.packages(pkgs = "LearnBayes", repos = "http://cran.us.r-project.org")

library(LearnBayes)
```

Note: The argument `repos =` in the `install.packages()` function directs where the package is obtained on CRAN (comprehensive R archive network). The CRAN repository is set automatically when using RStudio and you can install packages by clicking on *Packages* > *Install* in the lower-right panel

You need to specify the repository. When you render the markdown file (using the `Knit` button)  you don't want to install packages that already exist on your computer so you add the conditional `if()` function that says "only install the package IF it is not (`!`) available"

Make a copy of the data frame by assigning it to an object with the name `df` and print the first six rows using the `head()` function

```{r}
df <- studentdata
head(df)
```

Data frames are like spreadsheets with rows and columns. Each row is an observation (here each row is a student in an intro stats class at Bowling Green State University) and each column is a variable. Here the variables are answers to questions like what is your height, choose a number between 1 and 10, what time did you go to bed last night, etc

The names of the columns are printed using the `names()` function

```{r}
names(df)
```

All columns are of the same length, but not all students answered all questions so some of the data frame cells contain the missing-value indicator `NA` (not available)

Values in a data frame stored in rows and columns and are accessed with bracket notation [row, column] where row is the row number and column is the column number like a matrix

For example here you specify the value in the 10th row and 2nd column (`Height` column) of the `df` data frame

```{r}
df[10, 2]
```

By specifying only the row index and leaving the column index blank you get all values in that row, which corresponds to all the responses given by the 10th student

```{r}
df[10, ]
```

Drink preference was one of the questions. Responses across all students are available in the column labeled `Drink` as a vector of character values. You list the various drink preferences by typing

```{r}
df$Drink
```

Some students left that response blank and therefore the response is coded with the missing-value indicator

The variable type depends on the question asked. For example, answers given to the question of student height result in a numeric variable, answers given to the question about drink preference result in a character (or factor) variable

For integer, character, and factor variables you summarize the set of responses with the `table()` function

```{r}
table(df$Drink)
```

There are 113 students who prefer milk, 178 prefer soda, and 355 prefer water

Here you use the`plot()` method to make a draft plot of this table

```{r}
plot(x = df$Drink)
```

Notice that the sum of the responses is `r sum(table(df$Drink))`, which is less than the total number of students (`r nrow(df)`).

Students who left that question blank are ignored in the `table()` function. To include the missing values you add the argument `useNA = "ifany"` to the `table()` function.

```{r}
table(df$Drink, 
      useNA = "ifany")
```

Note: When you want code executed directly within the text you separate the code using single back ticks. This is useful when you write reports that need periodic updates when new data becomes available. Instead if you hard code the values in the text then you need to search the document for these values during each update of the report

Suppose you are interested in examining how long students reported sleeping during the night. This was not asked directly. You compute it from the `ToSleep` and `WakeUp` times columns. You assign the result of the difference to a column we call `SleepHrs`

```{r}
df$SleepHrs <- df$WakeUp - df$ToSleep
head(df)
```

Now you have a new numeric variable in the data frame called `SleepHrs`

You can't table numeric variables, but the `summary()` method prints a set of summary statistics for the set of values

```{r}
summary(df$SleepHrs)
```

The average number of hours slept is 7.4 with a maximum of 12.5 and a minimum of 2.5. There are four students that did not answer either when they went to sleep or when they woke up questions

You use the `hist()` function to plot a histogram of sleep hours. The function takes as input a vector of values

```{r}
hist(df$SleepHrs)
```

The function divides the number of sleep hours into one-hour bins and counts the number of students whose reported sleep hours falls into each bin. For example based on when they said they went to sleep and when the said they woke up, about 100 students slept between five and six hours the night before the survey

Is the amount of sleep for a student related to when they go to bed? If you place numeric variables on the x and y axes then you get a scatter plot

```{r}
plot(x = df$ToSleep,
     y = df$SleepHrs)
```

The `ToSleep` variable is centered on midnight so that -2 means a student went to sleep at 10p

You describe the decreasing relationship with a line through the points. The least-squares line is fit using the `lm()` function and the line is drawn on the existing plot with the `abline()` function applied to the linear regression object `model`.

```{r}
model <- lm(SleepHrs ~ ToSleep, 
            data = df)

plot(x = df$ToSleep,
     y = df$SleepHrs)
abline(model)
```

## Getting data into R {-}

You often start a research project by getting data into R. The file is on your computer or in the cloud. Secondary source data should be imported directly from repositories on the Web. When there is no direct link or API (application programming interface) to the repository, you need to first download the data

For example, consider the annually updated reports of tornadoes in the United States. The data repository is the Storm Prediction Center (SPC) <https://www.spc.noaa.gov/wcm/index.html#data>

Here you are interested in the file `1950-2021_actual_tornadoes.csv`. First download the file from the site with the `download.file()` function specifying the location (`url =`) and a name you want the file to be called on your computer (`destfile =`)

```{r}
download.file(url = "http://www.spc.noaa.gov/wcm/data/1950-2021_actual_tornadoes.csv",
              destfile = here::here("data", "Tornadoes.csv"))
```

A file called `Tornadoes.csv` is now be located in the directory `data`. Click on the *Files* tab in the lower-right panel, then select the `data` folder

Next you read (import) the file as a data frame using the `readr::read_csv()` function from the {tidyverse} group of packages.

```{r}
Torn.df <- readr::read_csv(file = here::here("data", "Tornadoes.csv"))
```

You preview the data frame using the `head()` function

```{r}
head(Torn.df)
```

Each row is a unique tornado report. Observations for each report include variables like the day and time, the state (`st`), the maximum EF rating (`mag`), the number of injuries (`inj`), the number of fatalities (`fat`), estimated property losses (`loss`), estimated crop losses (`closs`), start and end locations in decimal degrees longitude and latitude, length of the damage path in miles (`len`), width of the damage in yards (`wid`)

The total number of tornado reports in the data set is returned using the `nrow()` function

```{r}
nrow(Torn.df)
```

To create a subset of the data frame that contains only tornadoes in years (`yr`) since 2001, you include the logical operator `yr >= 2001` inside the subset operator (`[]`). The logical operator is placed in front of the comma since you want all *rows* where the result of the operator returns a value `TRUE`. After the column is left blank since you want all the variables

```{r}
Torn2.df <- Torn.df[Torn.df$yr >= 2001, ]
```

You see that there are fewer rows (tornado reports) in this new data frame assigned the object name `Torn2.df`.

You subset again, keeping only tornadoes with EF ratings (`mag` variable) greater than zero. Here you *recycle* the name `Torn2.df`

```{r}
Torn2.df <- Torn2.df[Torn2.df$mag > 0, ]
```

Now you compute the correlation between EF rating (`mag`) and path length (`len`) with the `cor()` function. The first argument is the vector of EF ratings and the second argument is the vector of path lengths

```{r}
cor(Torn2.df$mag, Torn2.df$len)
```

Path length is recorded in miles and path width in yards and the EF damage rating variable `mag` is numeric. Here you convert path length to kilometers, path width to meters, and the EF rating to a factor and include these changes as new columns in the data frame.

```{r}
Torn2.df$Length <- Torn2.df$len * 1609.34
Torn2.df$Width <- Torn2.df$wid * .9144
Torn2.df$EF <- factor(Torn2.df$mag)
```

Create side-by-side box plots of path length (in kilometers) by EF rating.

```{r}
plot(x = Torn2.df$EF, 
     y = Torn2.df$Length/1000)
```

Another example: Annual U.S. hurricane counts

Here you import the data directly from the Web by specifying the URL as a character string using the `file =` argument

```{r}
loc <- "http://myweb.fsu.edu/jelsner/temp/data/US.txt"
USHur.df <- readr::read_table(file = loc)
```

The `dim()` function returns the size of the data frame defined as the number of rows and the number of columns

```{r}
dim(USHur.df)
```

There are 166 rows and 6 columns in the data frame. Each row is a year and the columns include `Year`, number of hurricanes (`All`), number of major hurricanes (`MUS`), number of Gulf coast hurricanes (`G`), number of Florida hurricanes (`FL`), and number of East coast hurricanes (`E`) in that order

To get a glimpse of the data values you list the first six lines of the data frame using the `head()` function

```{r}
head(USHur.df)
```

The distribution of Florida hurricane counts by year is obtained using the `table()` function and specifying the `FL` column with `df$FL`

```{r}
table(USHur.df$FL)
```

There are 93 years without a FL hurricane, 43 years with exactly one hurricane, 24 years with two hurricanes, and so on

Another example: Statewide average rainfall in Florida. Work through it on your own

The data are monthly statewide average rainfall (in inches) for Florida starting in 1895. The data were obtained from <http://www.esrl.noaa.gov/psd/data/timeseries/>

I put values into a text editor and then uploaded the file to <http://myweb.fsu.edu/jelsner/temp/data/FLprecip.txt>

You get the data into R with the `readr::read_table()` function. You assign the data object the name `FLp.df`. You type the name of the object to see that it is a tabled data frame (tibble) with 117 rows and 13 columns

```{r}
loc <- "http://myweb.fsu.edu/jelsner/temp/data/FLprecip.txt"
FLp.df <- readr::read_table(file = loc)
FLp.df
```

The first column is the year and the next 12 columns are the months.

What was the statewide rainfall during June of 1900?

What happens when you check the column (vector) labeled `Year` to see what elements are equal to 1900?

```{r}
FLp.df$Year == 1900
```

Returns a vector with one element equal to `TRUE`.

What happens when you use this vector to select the vector of June rainfall values?

```{r}
FLp.df$Jun[FLp.df$Year == 1900]
```

What year had the wettest March?

```{r}
FLp.df$Mar
max(FLp.df$Mar)
which.max(FLp.df$Mar)
FLp.df$Year[which.max(FLp.df$Mar)]
```

What month during 1965 was the wettest? How wet was it?

```{r}
FLp.df$Year == 1965
FLp.df[FLp.df$Year == 1965, ]
which.max(FLp.df[FLp.df$Year == 1965, 2:12])
which.max(FLp.df[FLp.df$Year == 1965, 2:12])
max(FLp.df[FLp.df$Year == 1965, 2:12])
```

## Working with data frames using functions from {dplyr} {-}

The functions in the {dplyr} package (part of the tidyverse set of packages) simplify working with data frames. The functions work only on data frames

Function names are English language *verbs* so they are easier to remember. The verbs help you translate thought into code

Here we consider some of verbs using the `airquality` data frame. The data frame contains air quality measurements taken in New York City between May and September 1973 (`?airquality`) at time when pollution levels in the city were at an all time high

```{r}
dim(airquality)
head(airquality)
```

The columns include `Ozone` (ozone concentration in ppb), `Solar.R` (solar radiation in langleys), `Wind` (wind speed in mph), `Temp` (air temperature in degrees F), `Month`, and `Day`

You get summary statistics on the values in each column with the `summary()` method

```{r}
summary(airquality)
```

Note that columns that have missing values are tabulated. For example, there are 37 missing ozone measurements and 7 missing radiation measurements

Importantly for making your code more human readable you can apply the `summary()` function on the `airquality` data frame using the pipe operator (`|>`)

```{r}
airquality |> summary()
```

You read the pipe as THEN. "take the `airquality` data frame THEN summarize the columns"

The pipe operator allows you to string together functions that when read by a human makes it easy to understand what is being done

Think of it this way: suppose the object of interest is called `me` and there is a function called `wake_up()`. I could apply this function called `wake_up()` in two ways

```{r, eval=FALSE}
wake_up(me)  # classic

me |> wake_up()  # tidyverse
```

The second way involves more typing but it is easier to read (the subject comes before the predicate like simple English grammar) and thus easier to interpret. This becomes clear when stringing many functions together

Continuing, what happens to the result of `me` after the function `wake_up()` has been applied? I `get_out_of_bed()` and then `get_dressed()`

Again, you can apply these functions in two ways

```{r, eval=FALSE}
get_dressed(get_out_of_bed(wake_up(me))) # classic

me |> # tidyverse
  wake_up() |>
  get_out_of_bed() |>
  get_dressed()
```

The order of the functions often matters. I can't get dressed before I wake up but I could get dressed without getting out of bed

Note how I format the code. Each line is gets only one verb and each line ends with the pipe (except the last one). This makes it easy to read. Continuing

```{r, eval=FALSE}
me |>
  wake_up() |>
  get_out_of_bed() |>
  get_dressed() |>
  make_coffee() |>
  drink_coffee() |>
  leave_house()
```

Which is much better in terms of 'readability' then `leave_house(drink_coffee(make_coffee(get_dressed(get_out_of_bed(wake_up(me))))))`.

Now lets look at some of the commonly used verbs and to see how to apply them to a data frame

The function `select()` chooses variables by name. For example, choose the month (`Month`), day (`Day`), and temperature (`Temp`) columns.

```{r}
airquality |>
  dplyr::select(Month, Day, Temp)
```

The result is a data frame containing only the three columns with column names listed in the `select()` function.

Suppose you want a new data frame with only the temperature and ozone concentrations. You include an assignment operator (`<-`) and an object name (here `df`).

```{r}
df <- airquality |>
        dplyr::select(Temp, Ozone)
df
```

The verbs take data frames as input and return data frames

The function `filter()` chooses observations based on specific values. Suppose you want only the observations where the temperature is at or above 80 F

```{r}
airquality |>
  dplyr::filter(Temp >= 80)
```

The result is a data frame with the same 6 columns but now only 73 observations. Each of the observations has a temperature of at least 80 F

Suppose you want a new data frame keeping only observations when temperature is at least 80 F and when winds are less than 5 mph

```{r}
df <- airquality |> 
  dplyr::filter(Temp >= 80 & Wind < 5)
df
```

The function `arrange()` orders the rows by values given in a particular column.

```{r}
airquality |>
  dplyr::arrange(Solar.R)
```

The ordering is done from the lowest value of radiation to highest value. Here you see the first 10 rows. Note `Month` and `Day` are no longer chronological

Repeat, but order by the value of air temperature

```{r}
airquality |>
  dplyr::arrange(Temp)
```

Importantly you can string the functions together. For example select the variables radiation, wind, and temperature then filter by temperatures above 90 F and arrange by temperature

```{r}
airquality |>
  dplyr::select(Solar.R, Wind, Temp) |>
  dplyr::filter(Temp > 90) |>
  dplyr::arrange(Temp)
```

The result is a data frame with three columns and 14 rows arranged by increasing temperatures above 90 F

The `mutate()` function adds new columns to the data frame. For example, create a new column called `TempC` as the temperature in degrees Celsius. Also create a column called `WindMS` as the wind speed in meters per second

```{r}
airquality |>
  dplyr::mutate(TempC = (Temp - 32) * 5/9,
                WindMS = Wind * .44704) 
```

The resulting data frame has 8 columns (two new ones) labeled `TempC` and `WindMS`

On days when the temperature is below 60 F add a column giving the apparent temperature based on the cooling effect of the wind (wind chill) and then arrange from coldest to warmest apparent temperature

```{r}
airquality |>
  dplyr::filter(Temp < 60) |>
  dplyr::mutate(TempAp = 35.74 + .6215 * Temp - 35.75 * Wind^.16 + .4275 * Temp * Wind^.16) |>
  dplyr::arrange(TempAp)
```

The `summarize()` function reduces the data frame based on a function that computes a statistic. For example, to compute the average wind speed during July or the average temperature during June type

```{r}
airquality |>
  dplyr::filter(Month == 7) |>
  dplyr::summarize(Wavg = mean(Wind))

airquality |>
  dplyr::filter(Month == 6) |>
  dplyr::summarize(Tavg = mean(Temp))
```

You've seen functions that compute statistics including `sum()`, `sd()`, `min()`, `max()`, `var()`, `range()`, `median()`. Others include

|      Summary function | Description               |
|----------------------:|:--------------------------|
|          `dplyr::n()` | Length of the column      |
|      `dplyr::first()` | First value of the column |
|       `dplyr::last()` | Last value of the column  |
| `dplyr::n_distinct()` | Number of distinct values |

Find the maximum and median wind speed and maximum ozone concentration values during the month of May. Also determine the number of observations during May

```{r}
airquality |>
  dplyr::filter(Month == 5) |>
  dplyr::summarize(Wmax = max(Wind),
                   Wmed = median(Wind),
                   OzoneMax = max(Ozone),
                   NumDays = dplyr::n())
```

The result gives an `NA` for the maximum value of ozone (`OzoneMax`) because there is at least one missing value in the `Ozone` column. You fix this with the `na.rm = TRUE` argument in the function `max()`

```{r}
airquality |>
  dplyr::filter(Month == 5) |>
  dplyr::summarize(Wmax = max(Wind),
                   Wmed = median(Wind),
                   OzoneMax = max(Ozone, na.rm = TRUE),
                   NumDays = dplyr::n())
```

If you want to summarize separately for each month you use the `group_by()` function. You split the data frame by some variable (e.g., `Month`), apply a function to the individual data frames, and then combine the output

Find the highest ozone concentration by month. Include the number of observations (days) in the month

```{r}
airquality |>
  dplyr::group_by(Month) |>
  dplyr::summarize(OzoneMax =  max(Ozone, na.rm = TRUE),
                   NumDays = dplyr::n())
```

Find the average ozone concentration when temperatures are above and below 70 F. Include the number of observations (days) in the two groups

```{r}
airquality |>
  dplyr::group_by(Temp >= 70) |>
  dplyr::summarize(OzoneAvg =  mean(Ozone, na.rm = TRUE),
                   NumDays = dplyr::n())
```

On average ozone concentration is higher on warm days (Temp >= 70 F) days. Said another way; mean ozone concentration statistically depends on temperature

The mean is a model for the data. The statistical dependency of the mean implies that a model for ozone concentration will be improved by including temperature as an explanatory variable

In summary, the important verbs are

|          Verb | Description                                                                    |
|--------------------------------------:|:--------------------------------|
|    `select()` | selects columns; pick variables by their names                                 |
|    `filter()` | filters rows; pick observations by their values                                |
|   `arrange()` | re-orders the rows                                                             |
|    `mutate()` | creates new columns; create new variables with functions of existing variables |
| `summarize()` | summarizes values; collapse many values down to a single summary               |
|  `group_by()` | allows operations to be grouped                                                |

The six functions form the basis of a grammar for data. You can only alter a data frame by reordering the rows (`arrange()`), picking observations and variables of interest (`filter()` and `select()`), adding new variables that are functions of existing variables (`mutate()`), collapsing many values to a summary (`summarise()`), and conditioning on variables (`group_by()`)

The syntax of the functions are all the same

-   The first argument is a data frame. This argument is implicit when using the `|>` operator
-   The subsequent arguments describe what to do with the data frame. You refer to columns in the data frame directly (without using `$`)
-   The result is a new data frame

These properties make it easy to chain together many simple lines of code to do complex data manipulations and summaries all while making it easy to read by humans
