# Thursday February 2, 2023 {-}

**"Hell isn't other people's code. Hell is your own code from 3 years ago."** -- Jeff Atwood

* What we've covered thus far

| Lab | Date                  | Lessons covered |
|-----|-----------------------|-----------------|
| 1   | Thursday January 26   | 1-4             | *
| 2   | Thursday February 9   | 5-7             |
| 3   | Tuesday February 28   | 8-11            |
| 4   | Thursday April 6      | 12-18           |
| 5   | Thursday April 27     | 19-22           |

| Date     | Lesson/Lab|
|----------|-----------|
|January 10 | Syllabus | *
|January 12 | Lesson 1 | *
|January 17 | Lesson 2 | *
|January 19 | Lesson 3 | *
|January 24 | Lesson 4 | *
|January 26 | Lab 1 |    *
|January 31 | Lesson 5 | *
|February 2 | Lesson 6 | 
|February 7 | Lesson 7 |
|February 9 | Lab 2 |
|February 14 | Lesson 8 |
|February 16 | Lesson 9 |
|February 21 | Lesson 10 |
|February 23 | Lesson 11 |
|February 28 | Lab 3 |
|March 2 | Lesson 12 |
|March 7 | Lesson 13 |
|March 9 | Lesson 14 |
|March 14 | Spring Break |
|March 16 | Spring Break |
|March 21 | Lesson 15 |
|March 23 | FREE |
|March 28 | Lesson 16 |
|March 30 | Lesson 17 |
|April 4 | Lesson 18 |
|April 6 | Lab 4 |
|April 11 | Lesson 19 |
|April 13 | Lesson 20 |
|April 18 | Lesson 21 |
|April 20 | FREE |
|April 25 | Lesson 22 |
|April 27 | Lab 5 |

## Working with raster data {-}

The *raster data model* divides geographic space into a grid of cells (or pixels) of constant size (resolution). A cell may have an attribute value

The {raster} package has functions for working with raster data using S4 objects (reference class). The {terra} package has functions for creating, reading, manipulating, and writing raster data using S3 objects `SpatRaster` and `SpatVect`

To see what methods (functions) for class `SpatRaster` are available use the `methods()` function

```{r}
library(terra)
methods(class = "SpatRaster")
```

The list includes {base} R and {sf} functions and methods

The `terra::rast()` function defaults to a raster with a geographic (longitude/latitude) CRS and a 1 by 1 degree grid of cells across the globe

```{r}
( r <- terra::rast() )
```

The object has class `SpatRaster` with geographic coordinates spanning the globe at one-degree resolution in the north-south and the east-west directions

Arguments including `xmin`, `nrows`, `ncols`, and `crs` are used to change the default settings

To create a raster with 36 longitudes -100 and 0 degrees East longitude and 18 latitudes between the equator and 50 degrees N latitude you specify the number of columns, the number of rows and the extent as follows

```{r}
( r <- terra::rast(ncols = 36, nrows = 18, 
                 xmin = -100, xmax = 0, 
                 ymin = 0, ymax = 50) )
```

This results in raster with cell resolution of 2.7 degrees of longitude and 2.7 degrees of latitude

Extract only the resolution

```{r}
r |>
  terra::res()
```

The structure of the raster can be changed after created. Here you change the resolution to 3 degrees. This induces changes to the number of rows and columns

```{r}
terra::res(r) <- 3

r |>
  ncol()

r |>
  nrow()
```

The `SpatRaster` object `r` has no values assigned to the cells

```{r}
r <- terra::rast(ncol = 10, nrow = 10)

r |>
  terra::ncell()

r |>
  terra::hasValues()
```

Here there are 100 cells in a 10 by 10 arrangement with no values in any of the cells

The `terra::values()` function places values in the cells. The function is specified on the left-hand side of the assignment operator. Here you first assign to a vector of length `terra::ncell(r)` random numbers from a uniform distribution with the `runif()` function. The default is that the random numbers are between 0 and 1. Then you place those values in the grid cells

```{r}
v <- r |>
  terra::ncell() |>
  runif()

terra::values(r) <- v

r |>
  head()
```

The cells are arranged in lexicographical order (upper left to lower right) and the cells receive values from the vector in this order

The `terra::plot()` function creates a choropleth map of the values in cells

```{r}
r |>
 terra::plot()
```

The default CRS is geographic

```{r}
r |>
  terra::crs()
```

To re-project the raster use the function `terra::project()`

Here you create another raster with cell numbers as values using the `terra::setValues()` function to place the numbers in the cells. This function receives the values on the right-hand side of the assignment operator allowing you to use the pipe operator

```{r}
r <- terra::rast(xmin = -110, xmax = -90, 
                 ymin = 40, ymax = 60, 
                 ncols = 10, nrows = 10)

r <- r |>
  terra::setValues(1:terra::ncell(r))

r |>
  plot()
```

The values increase starting from top left to bottom right as dictated by the sequence `1:terra::ncell(r)` and the lexicographic order in which the raster grids are filled

The `terra::rast()` function imports data with functions from the {rgdal} package. Supported formats include `GeoTIFF`, `ESRI`, `ENVI`, and `ERDAS`. Formats that can be imported as a raster can also be used to export a raster

Consider the `Meuse` dataset (from the {sp} package), using a file in the native 'raster- file' format

```{r}
( r <- system.file("external/test.grd", 
                   package = "raster") |>
 terra::rast() )
```

Do the cells contain values? Is the raster stored in memory? Create a plot

```{r}
r |>
  terra::hasValues()

r |>
  terra::inMemory()

r |>
  terra::plot(main = "Raster layer from file")
```

Cells are always in a rectangular arrangement. Cells that lie outside a non-rectangular boundary inside the rectangular extent are set to `NA` and are not plotted

`SpatRaster` objects can have more than one attribute called layers

```{r}
r
```

The dimensions are `nrow` = 115 by `ncol` = 80 and `nlyr` = 1

Here you create three rasters and assign random values to the cells

```{r}
r1 <- terra::rast(nrow = 10, ncol = 10)
terra::values(r1) <- runif(terra::ncell(r1))
r2 <- terra::rast(nrow = 10, ncol = 10)
terra::values(r2) <- runif(terra::ncell(r2))
r3 <- terra::rast(nrow = 10, ncol = 10)
terra::values(r3) <- runif(terra::ncell(r3))
```

You combine the rasters into a single `SpatRaster` raster with three layers using the concatenate function `c()`

```{r}
( s <- c(r1, r2, r3) )
```

Extract dimensions and then plot

```{r}
s |>
  dim()

s |>
  terra::nlyr()

s |> 
  terra::plot()
```

Here you import a raster with multiple layers from a file. The layers are values corresponding to the RBG color scheme

```{r}
b <- system.file("external/rlogo.grd", 
                 package = "raster") |>
  terra::rast()

b |>
  terra::plot()
```

Most {base} R functions (`+`, `*`, `round()`, `ceiling()`, `log()`, etc) work on raster objects. Operations are done on each cell

Here you place the numbers from 1 to 100 sequentially in the cells, then add 100 to these values and take the square root

```{r}
r <- terra::rast(ncol = 10, nrow = 10)
terra::values(r) <- 1:terra::ncell(r)

s <- r + 100

s <- s |>
  sqrt()

s |>
  terra::plot()
```

Here you replace the cell values with random uniform numbers between 0 and 1. Then round to the nearest integer and add one

```{r}
r <- terra::rast(ncol = 10, nrow = 10)
terra::values(r) <- runif(terra::ncell(r))
r <- r |>
  round()

r <- r + 1

r |>
  terra::plot()
```

Replace only certain values with the subset function `[]`

```{r}
r <- terra::rast(xmin = -90, xmax = 90, ymin = -30, ymax = 30)
terra::values(r) <- rnorm(terra::ncell(r))
terra::plot(r)

r[r > 2] <- 0

r |>
  terra::plot()
```

Functions for manipulating a raster

The `terra::crop()` function takes a geographic subset of a larger raster object. A raster is cropped by providing an extent object or other spatial object from which an extent can be extracted (objects from classes deriving from raster and from spatial in the {sp} package)

The `terra::trim()` function crops a raster layer by removing the outer rows and columns that only contain `NA` values. The `terra::extend()` function adds new rows and/or columns with `NA` values

The `terra::merge()` function combines two or more rasters into a single raster. The input objects must have the same resolution and origin (such that their cells fit into a single larger raster). If this is not the case, first adjust one of the objects with the functions `aggregate()` or `resample()`

The `terra::aggregate()` and `terra::disagg()` functions change the resolution (cell size) of a raster object

As an example here you crop the raster into two pieces and then merge the two pieces into one. The `terra::merge()` function has the argument `filename =` that allows you to export to a file (here `test.grd` in the folder `outputs`)

```{r}
r1 <- r |>
  terra::crop(terra::ext(-180, 0, 0, 30)) 

r2 <- r |>
  terra::crop(terra::ext(-10, 180, -20, 10))

m <- r1 |>
  terra::merge(r2, 
               filename = here::here("data", "test.grd"), 
               overwrite = TRUE)
m |>
  plot()
```

The `terra::flip()` function flips the data (reverse order) in the horizontal or vertical direction. The `terra::rotate()` function rotates a raster that have longitudes from 0 to 360 degrees (often used by climatologists) to the standard -180 to 180 degrees system

You extract values from a raster for a set of locations with the `terra::extract()` function. The locations can be a vector object (points, lines, polygons), a matrix with (x, y) or (longitude, latitude -- in that order) coordinates, or a vector with cell numbers

First create a raster layer with integer values from 1 to 25 and specify two locations within the raster extent first as a matrix array and the second as a `SpatVector` object

```{r}
r <- terra::rast(ncols = 5, nrows = 5, 
                 xmin = 0, xmax = 5, 
                 ymin = 0, ymax = 5)
terra::values(r) <- 1:25

xy <- rbind(c(.5, .5), c(2.5, 2.5))

p <- xy |>
  terra::vect(crs = 'epsg:4326')
```

Next use the `terra::extract()` function with the first argument the raster layer and the second the locations from which you extract the values

```{r}
values1 <- r |>
  terra::extract(xy)

values2 <- r |>
  terra::extract(p)
```

In both cases the values are returned as a data frame. In the second case an `ID` variable is included

To convert the values of a raster layer to points or polygons you use `as.points()` and `as.polygons()`

```{r}
valuesAll <- r |>
  as.points()

```

These functions return a `SpatVector` object only for cells with values

```{r}
valuesAll |>
  plot()
```

Converting vector data to rasters

Vector data is converted to a raster with the `terra::rasterize()` function. Polygon to raster conversion is often done to create a mask (i.e. to set to `NA` a set of cells of a raster object, or to summarize values on a raster by zone. For example a country polygon is converted to a raster that is used to set all the cells outside that country to `NA`

Also polygons representing administrative regions such as states can be converted to a raster to summarize values by region. Point to raster conversion is often done to analyze location data (location of a specific species of tree in a forest)

Example: create a raster containing the number of tornadoes passing through each grid cell

Here you use a latitude/longitude grid (1/2 degree latitude by 1/2 degree longitude) with each cell in the grid containing the number of tornadoes that went through it since 2003

First import the tornado (initial track point) data as a simple feature data frame

```{r}
if(!"1950-2021-torn-initpoint" %in% list.files(here::here("data"))) {
download.file(url = "http://www.spc.noaa.gov/gis/svrgis/zipped/1950-2021-torn-initpoint.zip",
              destfile = here::here("data", "1950-2021-torn-initpoint.zip"))
unzip(here::here("data", "1950-2021-torn-initpoint.zip"), 
      exdir = here::here("data"))
}

Torn.sf <- sf::st_read(dsn = here::here("data", "1950-2021-torn-initpoint"), 
                       layer = "1950-2021-torn-initpoint") |>
  dplyr::filter(yr >= 2003)
```

Note the extent of the bounding box and check the CRS

```{r}
Torn.sf |>
  sf::st_crs()
```

Next create a raster (called `frame`) with a domain that covers the area of interest and assign a resolution of one degree in longitude and one degree in latitude. Check the extent of the raster with the `terra::ext()` function

```{r}
frame <- terra::rast(xmin = -106, xmax = -67, 
                     ymin = 24, ymax = 50)
terra::res(frame) <- .5

frame |>
  terra::ext()
```

Next use the `terra::rasterize()` function to count the number of times each raster cell contains a tornado. The first argument is the spatial data frame and the second is the raster without values. The argument `field =` specifies a column name in the spatial data frame (here just an identifier) and the argument `fun =` specifies what to do

Here you want a count of the unique instances of the field in each cell and this is done with setting `fun = "length"`. Raster cells without tornadoes are given a value of 0 based on the `background =` argument

```{r}
Torn.v <- Torn.sf |>
  terra::vect()

( Torn.r <- Torn.v |>
  terra::rasterize(y = frame, 
                   fun = "length",
                   background = 0) )
```

The result is a raster layer. The number of tornadoes occurring in each cell are the values

Print the first 200 values (lexicographical order).

```{r}
terra::values(Torn.r)[1:200]
```

Use the `plot()` method to visualize the raster layer

```{r}
Torn.r |>
  plot()
```

You can recognize the broad shape of the eastern 2/3rds of the United States. Some cells across the Plains and the South have quite a few tornadoes and very few tornadoes in cells over the Appalachian Mountains

Indeed tornado activity appears in distinct clusters. A statistic that estimates the amount of clustering is called Moran's I. It is a global measure of clustering with high values indicated by high values nearby to other high values and low values nearby to other low values

Values of Moran's I range from -1 to +1 where positive values indicate clustering and negative values indicate regularity (e.g., chessboard). It is implemented on a raster with the `raster::Moran()` function

The function works only with S4 raster objects. So you need to first convert `Torn.r` from a `SpatRaster` to a `RasterLayer`. You do this with `raster()` function after loading the {raster} package

```{r}
Torn.r2 <- Torn.r |>
  raster::raster()
```

You check the structure of the object with the `str()` function
```{r}
Torn.r2 |>
  str()
```

The object `Torn.r2` is a `RasterLayer` as an S4 data class. Note the use of slots for storing the information

You can use the `raster::Moran()` function on the `RasterLayer` object

```{r}
Torn.r2 |>
  raster::Moran()
```

The value of .75 indicates high level of tornado clustering at the .5 degree lat/lon resolution

Under the null hypothesis of no spatial autocorrelation the expected value for Moran's I is close to zero [-1/(n-1), where n is the number of cells]

Clusters at a local level can be found using a local indicator of spatial autocorrelation. One such indicator is local Moran's I, which is computed at each cell (using the `MoranLocal()` function) so the result is a raster

```{r}
Torn_lmi.r <- Torn.r2 |>
  raster::MoranLocal()

Torn_lmi.r |>
  plot()
```

This type of plot makes is easy to identify the "hot" spots of tornadoes over parts of the South and the Central Plains

To convert the local Moran raster to a S4 spatial data frame with polygon geometries use the `rasterToPolygons()` function.

```{r}
( Torn_lmi.sp <- Torn_lmi.r |>
  raster::rasterToPolygons() )
```

Then convert the `SpatialPolygonsDataFrame` to a simple features data frame and make a plot

```{r}
Torn_lmi.sf <- Torn_lmi.sp |>
  sf::st_as_sf()

library(ggplot2)

ggplot(data = Torn_lmi.sf) +
  geom_sf(mapping = aes(fill = layer, color = layer))
```

Or using functions from the {tmap} package you map the raster layer directly.

```{r}
tmap::tmap_mode("view")

Torn_lmi.r |>
  tmap::tm_shape() +
    tmap::tm_raster(alpha = .7)
```

Focal (neighborhood) functions

The function `terra::focal()` computes statistics in a neighborhood of cells around a focal cell, putting the result in the focal cell of an output raster. The `terra::distance()` function computes the shortest distance to cells that are not `NA`. The `terra::direction()` function computes the direction towards (or from) the nearest cell that is not `NA`. The `terra::adjacent()` function determines which cells are adjacent to other cells

Consider a multi-band image taken from a Landsat 7 view of a small part of the Brazilian coast. It is included in the {stars} package and stored as a *GeoTIFF* file labeled `L7_ETMs.tif`. You import the image as a raster stack

```{r}
if(!require(stars)) install.packages("stars", 
                                     repos = "http://cran.us.r-project.org")
library(stars)

f <- system.file("tif/L7_ETMs.tif",
                  package = "stars")

( L7.rs <- raster::stack(f) )
```

The data `L7.rs` is a `RasterStack` object as a S4 reference class

You list the slot names and extract the extent and CRS using the `@` syntax.

```{r}
L7.rs@extent
L7.rs@crs
```

You extract a single band (layer) from the stack with the `layer =` argument in the `raster()` function. You then plot the raster values with the `plot()` method and compute the spatial autocorrelation with the `raster::Moran()` function

```{r}
L7.rB3 <- L7.rs |>
  raster::raster(layer = 3)


L7.rB3 |>
  plot()

L7.rB3 |>
  raster::Moran()
```

You convert the raster to an S3 reference class data frame with the `as.data.frame()` method. Here you do that and then compute the normalized difference vegetation index (NDVI) using columns `L7_ETMs.4` and `L7_ETMs.3` and the `mutate()` function from the {dplyr} package

NDVI indicates live green vegetation from satellite images. Higher values indicate more green vegetation, negative values indicate water.

```{r}
L7.df <- L7.rs |>
  as.data.frame() |>
  dplyr::mutate(NDVI = (L7_ETMs_4 - L7_ETMs_3)/(L7_ETMs_4 + L7_ETMs_3))
```

More examples and other functions for working with raster data using functions from the {terra} package are illustrated in <https://geocompr.robinlovelace.net/raster-vector.html>. Again, I encourage you to take a look at this valuable resource

## Working with space-time data {-}

Space-time data arrive in the form of multi-dimensional arrays. Examples include:

-   raster images
-   socio-economic or demographic data
-   environmental variables monitored at fixed stations
-   time series of satellite images with multiple spectral bands
-   spatial simulations
-   climate and weather model output

The {stars} package provides functions and methods for working with space-time data as multi-dimensional S3 reference class arrays.

To see what methods (functions) for class `stars` are available use the `methods()` function.

```{r}
methods(class = "stars")
```

The list includes {base} R and {tidyverse} methods.

The typical data array is that where two dimensions represent spatial raster dimensions and the third dimensions is a band (or time). [Data array](https://raw.githubusercontent.com/r-spatial/stars/master/images/cube1.png)

But arrays can have more dimensions. For example, time, space, spectral band, and sensor type. [Data cube](https://raw.githubusercontent.com/r-spatial/stars/master/images/cube2.png)

You import a set of rasters (raster stack) as a {stars} object using the `stars::read_stars()` function. Consider the multi-band image taken from a Landsat 7 view of a small part of the Brazilian coast. It is included in the {stars} package and stored as a *GeoTIFF* file labeled `L7_ETMs.tif`.

```{r}
( L7.stars <- system.file("tif/L7_ETMs.tif",
                        package = "stars") |>
  stars::read_stars() )
```

There are three dimensions to this {stars} object, two spatial (`x` and `y`), and the third across six bands (`band`). Values across the six bands and space are summarized as a single attribute with name `L7_ETMs.tif`

The data are stored in a four dimensional array. The first index is the attribute, the second and third indexes are the spatial coordinates, and the fourth index is the band

Here you plot bands 3 and 4 by sequencing on the fourth index and using the `plot()` method

```{r}
plot(L7.stars[,,,3:4])
```

Since the data object is S3 you use functions from the `ggplot2()` package together with the `geom_stars()` layer from the {stars} package to plot all 6 bands with a common color scale bar

```{r}
ggplot() +
  stars::geom_stars(data = L7.stars) +
  facet_wrap(~ band)
```

You create a new {stars} object by applying a function to the band values. For example here you compute normalized difference vegetation index (NDVI) through a function applied across the `x` and `y` spatial dimensions using the `stars::st_apply()` method after creating the function `NDVI()`

```{r}
NDVI <- function(z) (z[4] - z[3]) / (z[4] + z[3])

( NDVI.stars <- L7.stars |>
    stars::st_apply(MARGIN = c("x", "y"), 
                    FUN = NDVI) )
ggplot() +
  stars::geom_stars(data = NDVI.stars) 
```

The stars data frame can also be split, here on the band dimension, to yield a representation as six rasters in the list form

```{r}
( L7split.stars <- split(L7.stars, 
                         f = "band") )
```

Now the bands are given as columns in the data frame part of the {stars} object and there are only two dimensions (`x` and `y`)

Raster data do not need to be regular or aligned along the cardinal directions. Functions in the {stars} package supports rotated, sheared, rectilinear and curvi-linear grids
[Grids](https://raw.githubusercontent.com/r-spatial/stars/master/man/figures/README-plot2-1.png)

Functions in the {stars} package also support the vector data model. Vector data cubes arise when you have a single dimension that points to distinct spatial feature geometry, such as polygons (e.g. denoting administrative regions). [Vector data cube polygons](https://raw.githubusercontent.com/r-spatial/stars/master/images/cube3.png)

Or points (e.g., denoting sensor locations). [Vector data cube points](https://raw.githubusercontent.com/r-spatial/stars/master/images/cube4.png)

For more see: <https://github.com/r-spatial/stars/tree/master/vignettes> and <https://awesomeopensource.com/project/r-spatial/stars>

Also you can check out some rough code that I've been working on to take advantage of the {stars} functionality including plotting daily temperatures across the U.S. and creating a vector data cube of COVID19 data in the `stars.Rmd` file on course GitHub site in the folder `Other_Rmds`
