# Tuesday September 13, 2022 {.unnumbered}

**"Measuring programming progress by lines of code is like measuring aircraft building progress by weight."** – Bill Gates

## Today {.unnumbered}

- S4 spatial data objects
- Working with raster data

## S4 spatial data objects

The {sp} package also has methods for working with spatial data. A few of the packages we will use this semester for analyzing/modeling spatial data work only with {sp} objects so it is helpful to see how these objects are structured.

Install and load the package.
```{r}
if(!require(sp)) install.packages(pkgs = "sp", repos = "http://cran.us.r-project.org")

library(sp)
```

Spatial objects from the {sp} package fall into two types: 

1) spatial-only information (the geometry). Geometries include `SpatialPoints`, `SpatialLines`, `SpatialPolygons`, etc, and 
2) extensions to these types where attribute information is available and stored in a data frame. These include `SpatialPointsDataFrame`, `SpatialLinesDataFrame`, etc.

The typical situation is that you have a simple feature data frame (an S3 spatial object) that we've imported and we need to convert it to an {sp} spatial data frame (an S4 spatial object) before the information can be used to fit a model.

Consider again the the tornado tracks that you import as a simple feature data frame.
```{r}
FL_Torn.sf <- sf::st_read(dsn = here::here("data", "1950-2020-torn-aspath"), 
                          layer = "1950-2020-torn-aspath") |>
  dplyr::filter(st == "FL")
class(FL_Torn.sf)
```

The object `FL_Torn.sf` is a simple feature data frame. You convert the simple feature data frame to an S4 spatial object using the `sf::as_Spatial()` function.
```{r}
FL_Torn.sp <- FL_Torn.sf |>
  sf::as_Spatial()

class(FL_Torn.sp)
```

The file `FL_Torn.sp` is a S4 spatial object of class `SpatialLinesDataFrame`. 

Information in S4 spatial objects is stored in _slots_. Slot names are listed with the `slotNames()` function.
```{r}
FL_Torn.sp |>
  slotNames()
```

The `data` slot contains the data frame (attribute table), the `lines` slot contains the spatial geometries (in this case lines), the `bbox` slot is the boundary box and the `proj4string` slot is the CRS.

The object name followed by the `@` symbol allows access to information in the slot. The `@` symbol is similar to the `$` symbol for regular data frames. For example to see the first three rows of the data frame we type
```{r}
FL_Torn.sp@data[1:3, ]
```

You should recognize this as information about the first three tornadoes in the record. In fact, the object name together with the slot name `data` has class `data.frame`.
```{r}
class(FL_Torn.sp@data)
```

When using the `$` symbol on S4 spatial objects, you can access the data as a data frame. For example, to list the EF rating (column labeled `mag`) of the first 3 tornadoes.
```{r}
FL_Torn.sp$mag[1:3]
```

Selecting, retrieving, or replacing attributes in S4 spatial data frames is done with methods in {base} R package. For example `[]` is used to select rows and/or columns. To select `mag` of the 7th tornado type
```{r}
FL_Torn.sp$mag[7]
```

Other methods include: `plot()`, `summary()`,`dim()` and `names()` (operate on the data slot), `as.data.frame()`, `as.matrix()` and `image()` (for spatial data on a grid), and `length()` (number of features).

You can't use the {dplyr} verbs on S4 data frames. To convert from an S4 spatial data frame to a simple feature data frame, use `sf::st_as_sf()`.

The first spatial geometry is given as the first element of the lines list.
```{r}
FL_Torn.sp@lines[1]
```

It is an object of class `Lines`. The line is identified by a matrix indicating the longitude and latitude of the start point in row one and the longitude and latitude of the end point in row two.

The `bbox` slot is an object of class `matrix` and `array` and the `proj4string` slot is of class `CRS`.

The interface to the geometry engine-open source (GEOS) is through the {rgeos} package.

## Working with raster data

The _raster data model_ divides geographic space into a grid of cells of constant size (resolution) and we use classes from the {raster} package to work with raster data.

A raster is a data structure that divides space into rectangles called 'cells' (or 'pixels'). Each cell has an attribute value.

The {terra} package has functions for creating, reading, manipulating, and writing raster data and functions for raster manipulation common in GIS.

The `terra::rast()` function creates a raster with a geographic (longitude/latitude) CRS and a 1 by 1 degree grid of cells across the globe.
```{r}
r <- terra::rast()
r
```

Arguments including `xmin`, `nrows`, `ncols`, and `crs` are used to change these default settings.

The object has class `SpatRaster` with geographic coordinates spanning the globe at one-degree resolution in the north-south and the east-west directions.

To create a raster with 36 longitudes -100 and 0 degrees East longitude and 18 latitudes between the equator and 50 degrees N latitude we specify the number of columns, the number of rows and the extent as follows.
```{r}
r <- terra::rast(ncols = 36, nrows = 18, 
                 xmin = -100, xmax = 0, 
                 ymin = 0, ymax = 50)
r
terra::res(r)
```

This results in raster with cell resolution of 2.7 degrees of longitude and 2.7 degrees of latitude.

The structure of the raster can be changed after it is created. Here you change the resolution to 3 degrees. This induces changes to the number of rows and columns.
```{r}
terra::res(r) <- 3
ncol(r)
nrow(r)
```

The `SpatRaster` object `r` is a template with no values assigned to the cells and by default it will have an extent that spans the globe.
```{r}
r <- terra::rast(ncol = 10, nrow = 10)
terra::ncell(r)
terra::hasValues(r)
```

Here there are 100 cells in a 10 by 10 arrangement with no values in any of the cells.

The `terra::values()` function is used to place values in the cells. The function is specified on the left-hand side of the assignment operator. First you assign to a vector of length `terra::ncell(r)` random numbers from a uniform distribution with the `runif()` function. The default is that the random numbers are between 0 and 1.
```{r}
v <- runif(terra::ncell(r))
head(v)
terra::values(r) <- v
head(r)
```

The cells are arranged in lexicographical order (upper left to lower right) and the cells are populated with values from the vector in this order.

The `terra::plot()` function creates a choropleth map of the values in cells.
```{r}
terra::plot(r)
```

The default CRS is geographic.
```{r}
terra::crs(r)
```

To re-project the raster use the function `terra::project()`.

Here you create a new raster with cell numbers as values using the `terra::setValues()` function to place the numbers in the cells.
```{r}
r <- terra::rast(xmin = -110, xmax = -90, 
                 ymin = 40, ymax = 60, 
                 ncols = 10, nrows = 10)
r <- terra::setValues(r, 1:terra::ncell(r))
terra::plot(r)
```

The values increase starting from top left to bottom right as dictated by the sequence `1:terra::ncell(r)` and the lexicographic order in which the raster grids are filled.

The `terra::rast()` function imports data with functions from the {rgdal} package. Supported formats include `GeoTIFF`, `ESRI`, `ENVI`, and `ERDAS`. Most formats that can import a raster can also be used to export a raster. 

Consider the `Meuse` dataset (from the {sp} package), using a file in the native 'raster- file' format.
```{r}
f <- system.file("external/test.grd", 
                 package = "raster")
r <- terra::rast(f)
```

Do the cells contain values? Is the raster stored in memory? Create a plot.
```{r}
terra::hasValues(r)
terra::inMemory(r)
terra::plot(r, main = "Raster layer from file")
```

Note the raster is a set of cells arranged in a rectangular array. Values that are coded as `NA` are not plotted.

`SpatRaster` objects can have more than one raster. These are called layers.
```{r}
r
```

The dimensions are nrow = 115 by ncol = 80 and nlyr = 1.

You can add layers to the object. Here you create three rasters and assign random values to the cells.
```{r}
r1 <- terra::rast(nrow = 10, ncol = 10)
terra::values(r1) <- runif(terra::ncell(r1))
r2 <- terra::rast(nrow = 10, ncol = 10)
terra::values(r2) <- runif(terra::ncell(r2))
r3 <- terra::rast(nrow = 10, ncol = 10)
terra::values(r3) <- runif(terra::ncell(r3))
```

You combine the rasters into a single `SpatRaster` object with the concatenate function `c()`.
```{r}
s <- c(r1, r2, r3)
s
dim(s)
terra::nlyr(s)
terra::plot(s)
```

Each raster is a separate layer.

Here you import a set of raster layers from a file.
```{r}
f <- system.file("external/rlogo.grd", 
                 package = "raster")
b <- terra::rast(f)
b
terra::plot(b)
```

Most {base} R functions (`+`, `*`, `round()`, `ceiling()`, `log()`, etc) work on raster objects. Operations are done on all cells at once.

Here you place the numbers from 1 to 100 sequentially in the cells, then add 100 to these values and take the square root.
```{r}
r <- terra::rast(ncol = 10, nrow = 10)
terra::values(r) <- 1:terra::ncell(r)
s <- r + 100
s <- sqrt(s)
terra::plot(s)
```

Here you replace the cell values with random uniform numbers between 0 and 1. Then round to the nearest integer and add one.
```{r}
r <- terra::rast(ncol = 10, nrow = 10)
terra::values(r) <- runif(terra::ncell(r))
r <- round(r)
r <- r + 1
terra::plot(r)
```

Replace only certain values with the subset function `[]`.
```{r}
r <- terra::rast(xmin = -90, xmax = 90, ymin = -30, ymax = 30)
terra::plot(r)
r[r > 2] <- 0
terra::plot(r)
```

Functions for manipulating a raster

The `terra::crop()` function takes a geographic subset of a larger raster object. A raster is cropped by providing an extent object or other spatial object from which an extent can be extracted (objects from classes deriving from raster and from spatial in the {sp} package). 

The `terra::trim()` function crops a raster layer by removing the outer rows and columns that only contain `NA` values. The `terra::extend()` function adds new rows and/or columns with `NA` values.

The `terra::merge()` function combines two or more rasters into a single raster. The input objects must have the same resolution and origin (such that their cells fit into a single larger raster). If this is not the case, first adjust one of the objects with the functions `aggregate()` or `resample()`.

The `terra::aggregate()` and `terra::disagg()` functions change the resolution (cell size) of a raster object.

As a simple example showing some of this functionality here you crop the raster into two pieces and then merge the two pieces into one. The `terra::merge()` function has an argument that allows you to export to a file (here `test.grd`).
```{r}
r1 <- terra::crop(r, terra::ext(-180, 0, 0, 30)) 
r2 <- terra::crop(r, terra::ext(-10, 180, -20, 10))
m <- terra::merge(r1, r2, 
                  filename = 'test.grd', 
                  overwrite = TRUE)
terra::plot(m)
```

The `terra::flip()` function flips the data (reverse order) in the horizontal or vertical direction. The `terra::rotate()` function rotates a raster that have longitudes from 0 to 360 degrees (often used by climatologists) to the standard -180 to 180 degrees system.

You can extract values from a raster for a set of locations with the `terra::extract()` function. The locations can be a vector object (points, lines, polygons), a matrix with (x, y) or (longitude, latitude -- in that order!) coordinates, or a vector with cell numbers.

```{r}
r <- terra::rast(ncols = 5, nrows = 5, 
                 xmin = 0, xmax = 5, 
                 ymin = 0, ymax = 5)
values(r) <- 1:25

xy <- rbind(c(.5, .5), c(2.5, 2.5))
p <- vect(xy, crs="+proj=longlat +datum=WGS84")

terra::extract(r, xy)
terra::extract(r, p)
```

To convert the values of a raster layer to points or polygons we use `as.points()` and `as.polygons()`. These functions return a `SpatVector` object for cells that are not missing value.

Vector data is converted to a raster with the `terra::rasterize()` function. Polygon to raster conversion is often done to create a mask (i.e. to set to `NA` a set of cells of a raster object, or to summarize values on a raster by zone. For example a country polygon is converted to a raster that is used to set all the cells outside that country to `NA`. Also polygons representing administrative regions such as states can be converted to a raster to summarize values by region. Point to raster conversion is often done to analyze location data (location of a specific species of tree in a forest).

Example: the number of tornadoes passing through each grid cell

Here we want a latitude/longitude grid (1/2 degree latitude by 1/2 degree longitude) with each cell in the grid containing the number of tornadoes that went through it since 2003.

First import the tornado (initial track point) data as a simple feature data frame.
```{r}
if(!"1950-2020-torn-initpoint" %in% list.files(here::here("data"))) {
download.file(url = "http://www.spc.noaa.gov/gis/svrgis/zipped/1950-2020-torn-initpoint.zip",
              destfile = here::here("data", "1950-2020-torn-initpoint.zip"))
unzip(here::here("data", "1950-2020-torn-initpoint.zip"), 
      exdir = here::here("data"))
}
Torn.sf <- sf::st_read(dsn = here::here("data", "1950-2020-torn-initpoint"), 
                       layer = "1950-2020-torn-initpoint") |>
  dplyr::filter(yr >= 2003)
```

Note the extent of the bounding box and check the native CRS.
```{r}
sf::st_crs(Torn.sf)
```

The CRS is geographic.

Next create a raster (called `frame`) with a domain that covers the area of interest and assign a resolution of one degree in longitude and one degree in latitude. Check the extent of the raster with the `terra::ext()` function.
```{r}
frame <- terra::rast(xmin = -106, xmax = -67, 
                     ymin = 24, ymax = 50)
terra::res(frame) <- .5
terra::ext(frame)
```

Next use the `terra::rasterize()` function to count the number of times each raster cell contains a tornado. The first argument is the spatial data frame and the second is the raster without values. The argument `field =` specifies a column name in the spatial data frame (here just an identifier) and the argument `fun =` specifies what to do (here simply count the unique instances of the field in each cell). Raster cells without tornadoes are given a value of 0 based on the `background =` argument.
```{r}
Torn.v <- terra::vect(Torn.sf)

Torn.r <- terra::rasterize(x = terra::vect(Torn.sf), 
                           y = frame, 
                           field = "om", 
                           fun = "count",
                           background = 0)
class(Torn.r)
dim(Torn.r)
```

The result is a raster layer. The number of tornadoes occurring in each cell are the values.

We print out the first 200 values (lexicographical order).
```{r}
terra::values(Torn.r)[1:200]
```

To visualize the raster we use the `plot()` method.
```{r}
plot(Torn.r)
```

We can recognize the broad shape of the eastern 2/3rds of the United States. Some cells across the Plains and the South have quite a few tornadoes.

A spatial statistic indicating how similar values in neighboring cells tend to be is called Moran I. It is a global measure of clustering (high values near high values and low values near low values). Values range from -1 to +1 where positive values indicate clustering and negative values indicate regularity (e.g., chessboard). It is implemented on a raster with the `Moran()` function. 
```{r}
Moran(Torn.r)
```

The value of .74 indicates a high level of tornado clustering at this scale. Under the null hypothesis of no spatial autocorrelation the expected value for Moran I is close to zero [-1/(n-1), where n is the number of cells].

With raster data the neighborhood definition is constant across the study region.

Clusters at a local level can be found using a local indicator of spatial autocorrelation. One such indicator is local Moran I, which is computed at each cell (using the `MoranLocal()` function) so the result is a raster.
```{r}
Torn_lmi.r <- MoranLocal(Torn.r)
plot(Torn_lmi.r)
```

This type of plot makes is easier to see the hot spots of tornadoes over parts of the South and the Central Plains.

To convert the tornado raster to a S4 spatial data frame with polygon geometries we use the `rasterToPolygons()` function. 
```{r}
TornCounts.sp <- rasterToPolygons(Torn.r)
class(TornCounts.sp)
```

Then we convert the `SpatialPolygonsDataFrame` to a simple features data frame and make a plot.
```{r}
TornCounts.sf <- st_as_sf(TornCounts.sp)

ggplot(data = TornCounts.sf) +
  geom_sf(aes(fill = layer))
```

Focal (neighborhood) functions

The functions: `focal()`, `focalFilter()`, `focalNA()` compute statistics in a neighborhood of cells around a focal cell, putting the result in the focal cell of an output raster. 

With `focal()`, the neighborhood can only be a rectangle. With `focalFilter()`, the neighborhood is a user-defined a matrix of weights and could approximate any shape by giving some cells zero weight. The `focalNA()` function only computes new values for cells that are `NA` in the input raster.

The `distance()` function computes the shortest distance to cells that are not NA. The `pointDistance()` function computes the shortest distance to any point in a set of points. 

The `gridDistance()` function computes the distance when following grid cells that can be traversed (e.g. excluding water bodies). The `direction()` function computes the direction towards (or from) the nearest cell that is not `NA`. The `adjacency()` function determines which cells are adjacent to other cells, and the `pointDistance()` function computes distance between points.

Finally, you can coerce `SpatRasters` to raster objects, after loading the {raster} package, with `as(object, "Raster")`, or `raster(object)` or `brick(object)` or `stack(object)`
