# Thursday October 27, 2022 {.unnumbered}

**"Machines take me by surprise with great frequency."** -- Alan Turing

## Modeling the occurrence of violent tornadoes

The long-term risk of a tornado is estimated using historical records. The rarity of the most violent tornadoes makes the risk estimates for these most destructive events unstable. Here we use point pattern data models together with the more stable risk estimates from the larger set of less violent tornadoes to create more reliable estimates of the risk of a violent tornado.

We restrict our attention to tornadoes occurring across the state of Kansas during the period 1954--2018.

```{r}
library(sf)
library(tidyverse)

Torn.sf <- st_read(dsn = "1950-2018-torn-initpoint") %>%
  st_transform(crs = 3082) %>%
  filter(mag >= 0, yr >= 1954) %>%
  mutate(EF = mag,
         EFf = as.factor(EF)) %>%
  dplyr::select(yr, EF, EFf)

library(USAboundaries)
library(maptools)

W.sfc <- us_states(states = "Kansas") %>%
  st_transform(crs = st_crs(Torn.sf)) %>%
  st_geometry()

Torn.sf <- Torn.sf[W.sfc, ]
```

Create a `owin` and `ppp` objects. Note: we've already subset by Kansas tornadoes above but we need to subset on the `ppp` object to assign the KS boundary as the analysis window. Keep this in mind: Point pattern analysis and modeling are done with a set of event locations occurring within a boundary. If the boundary is not defined then a bounding box is used.

```{r}
library(spatstat)

KS.win <- W.sfc %>%
  as_Spatial() %>%
  as.owin()

T.ppp <- Torn.sf["EF"] %>%
  as_Spatial() %>%
  as.ppp()

T.ppp <- T.ppp[KS.win]

summary(T.ppp)
```

There are 4079 tornadoes over the period with an average intensity of 189 per 100 square kilometer (multiply the average intensity in square meters by 10\^10).

We separate the point pattern data into non-violent tornadoes and violent tornadoes. Non-violent tornadoes include those with an EF damage rating of 0, 1, 2 or 3. Violent tornadoes include those with an EF rating of 4 or 5. Note: Even non-violent tornadoes are capable of considerable damage and casualties.

```{r}
NV.ppp <- T.ppp %>%
  subset(marks <= 3 & marks >= 0) %>%
  unmark()

summary(NV.ppp)

V.ppp <- T.ppp %>%
  subset(marks >= 4) %>% 
  unmark()

summary(V.ppp)
```

The spatial intensity of the non-violent tornadoes is 187 per 100 sq km. The spatial intensity of the violent tornadoes is 1.8 per 100 square kilometer.

Locations of the violent tornadoes.

```{r}
plot(V.ppp)
```

In Lesson 16 we quantified the spatial intensity tornado reports as a function of distance to nearest city. We found there was a statistically significant relationship so we include this as an explanatory variable. Import the data, set the CRS, and transform the CRS to match that of the tornadoes. Exclude cities with fewer than 1000 people.

```{r}
C.sf <- us_cities() %>%
  filter(population >= 1000) %>%
  st_transform(crs = st_crs(Torn.sf))
```

Then create a `ppp` object of events from the city/town locations. First create an S4 class spatial data frame and convert this data frame to a `ppp` object before removing the marks. Then subset the events by the window.

```{r}
C.ppp <- C.sf %>%
  as_Spatial() %>%
  as.ppp()

C.ppp <- C.ppp[KS.win] %>%
  unmark()

plot(C.ppp)
```

Next create a distance map of the city/town locations using the `distmap()` function.

```{r}
Zc <- distmap(C.ppp)
plot(Zc)
```

The resulting object an `im` class indicating a pixel image. Pixel values are distances is meters. Blue indicates locations that are less than 20 km from a city or town with a population of at least 1000.

Our interest lies with the distance to nearest non-violent tornado. We check to see if this might be a useful variable in a model.

```{r}
Znv <- distmap(NV.ppp)
rhat <- rhohat(V.ppp, 
               covariate = Znv, 
               adjust = 1.5, 
               smoother = "kernel", 
               method = "transform")

dist <- rhat$Znv
rho <- rhat$rho
hi <- rhat$hi
lo <- rhat$lo
Rho.df <- data.frame(dist = dist, rho = rho, hi = hi, lo = lo)

ggplot(Rho.df) + 
  geom_ribbon(aes(x = dist, ymin = lo, ymax = hi), alpha = .3) + 
  geom_line(aes(x = dist, y = rho), col = "black") + 
  ylab("Spatial intensity of violent tornadoes") + xlab("Distance from nearest non-violent tornado (m)") + 
  theme_minimal()
```

This shows that regions that get non-violent tornadoes also see higher rates of violent tornadoes.

So our model includes two trend terms.

```{r}
model <- ppm(V.ppp, 
             trend = ~ Zc + Znv, 
             covariates = list(Zc = Zc, Znv = Znv))

model %>%
  summary() %>%
  coef()
```

As expected the model shows fewer violent tornadoes with increasing distance from the nearest city (negative coefficient on `Zc`) and fewer violent tornadoes with increasing distance from a non-violent tornado (negative coefficient on `Znv`).

Since the spatial unit is meters the coefficient of -3.06e-05 is interpreted as a [1 - exp(-.0306)] \* 100% or 3% decrease in violent tornado reports per kilometer of distance from a city. Similarly the coefficient on distance from nearest non-violent tornado is interpreted as a 23% decrease in violent tornado reports per kilometer of distance from nearest non-violent tornado.

We check if there is any residual nearest neighbor correlation.

```{r}
E <- envelope(model, 
              fun = Kest, 
              nsim = 39,
              global = TRUE)
plot(E, main = "Inhomogeneous Poisson Model", legend = FALSE)
```

There appears to be a bit of regularity at smaller scales. The empirical curve (black line) falls slightly below the model (dashed red line). There are fewer nearby violent tornadoes than one would expect.

To see if this is statistically significant, we add an inhibition process to the model.

```{r}
model2 <- ppm(V.ppp, 
              trend = ~ Zc + Znv, 
              covariates = list(Zc = Zc, Znv = Znv),
              interaction = Strauss(r = 40000))

model2 %>%
  summary() %>%
  coef()
```

We see a negative sign on the interaction coefficient as expected from the above plot, but the standard error is relatively large so it is not significant.

Remove the inhibition process and add a trend term in the east-west direction.

```{r}
model3 <- ppm(V.ppp, 
              trend = ~ Zc + Znv + x, 
              covariates = list(Zc = Zc, Znv = Znv))

model3 %>%
  summary() %>%
  coef()
```

There is a significant eastward trend but it appears to confound the distance to city term. Why is this?

Plot simulated data.

```{r}
plot(V.ppp)
plot(simulate(model, nsim = 6))
```

## Terrain roughness and tornado occurrence rates

Most thunderstorms fail to produce tornadoes. Tornado initiation is sensitive to an interplay of processes across a range of spatial scales, including the scale of a few hundred meters where the air flow near the ground is converging inward toward the tornado. It stands to reason that the underlying surface can affect this convergent inflow.

To make the analysis and modeling run faster we consider only Kansas tornadoes with an EF2 or higher rating.

```{r}
T.ppp25 <- T.ppp %>%
  subset(marks >= 2) %>%
  unmark()

summary(T.ppp25)
```

There are 626 tornadoes with an average intensity of 29 per 100 sq. km.

Plot the spatial varying intensity.

```{r}
T.ppp25 %>%
  density() %>%
  plot()
```

To quantify the relationship between distance to city and tornado report density we use a model.

```{r}
model0 <- ppm(T.ppp25, 
              trend = ~ Zc, 
              covariates = list(Zc = Zc))
model0
```

As expected the model shows a decreasing trend with increasing distance from cities (negative value on the `Zc` coefficient). The value is on the log scale so we do some arithmetic.

```{r}
100 * (1 - exp(coef(model0)[2] * 1000))
```

The coefficient is interpreted as a 1.7% decrease in the number of tornado reports per kilometer of distance from a city (on average).

Can we do better? Statistically: is the model adequate? Here we check model adequacy by examining model residuals against the assumption of homogeneous Poisson.

```{r}
E <- envelope(model0, 
              fun = Kest, 
              nsim = 39,
              global = TRUE)
plot(E, main = "", legend = FALSE)
```

We find that, after accounting for distance from nearest town, there is a tendency for tornado reports to cluster at all distances.

Elevation might be a factor in tornado occurrence rates. In particular the roughness of the underlying surface might make some areas more or less prone to tornadoes. Here we investigate this possibility using elevation data.

Digital elevation data are available from <http://www.viewfinderpanoramas.org>. The data has been uploaded to my website. Download and unzip the data.

```{r}
if(!"15-H.tif" %in% list.files()){
download.file(url = "http://myweb.fsu.edu/jelsner/temp/data/15-H_tif.zip",
              destfile = "15-H_tif.zip")
unzip("15-H_tif.zip")
}
```

Import the elevation raster and crop it to the extent of Kansas.

```{r}
library(raster)

Elev <- raster("15-H.tif")

W.sfc2 <- W.sfc %>%
  st_transform(crs = crs(Elev))

Elev2 <- crop(Elev, st_bbox(W.sfc2))
```

Map the elevation.

```{r}
library(tmap)

tm_shape(Elev2) +
  tm_raster() +
tm_shape(W.sfc2) +
  tm_borders() +
tm_layout(title = "Elevation (m)", 
          legend.outside = TRUE)
```

Get elevation roughness using the `terrain()` function from the {raster} package. Use the `projectRaster()` function to change the native projection to the projection of the tornado data. This takes a few seconds since the projection is not conformal and the grid spacing needs to be preserved.

```{r}
TR <- terrain(Elev2, opt = 'roughness')
ElevP <- projectRaster(Elev2, crs = st_crs(Torn.sf)$proj4string)
TRP <- projectRaster(TR, crs = st_crs(Torn.sf)$proj4string)

#writeRaster(ElevP, filename = "ElevP")
#writeRaster(TRP, filename = "TRP")
```

Create image objects from the elevation and roughness rasters. First convert the rasters to a S4 class spatial grid data frame.

```{r}
El <- ElevP %>%
  as("SpatialGridDataFrame") %>%
  as("im")
Tr <- TRP %>%
  as("SpatialGridDataFrame") %>%
  as("im")
```

Model the tornado events.

```{r}
model1 <- ppm(T.ppp25, 
             trend = ~ Zc + El + Tr, 
             covariates = list(Zc = Zc, El = El, Tr = Tr))
summary(model1)
```

We see that both elevation and terrain roughness are statistically significant in explaining the spatial pattern of tornadoes after accounting for distance to nearest city. Tornadoes are more common over lower elevations and smoother terrain.

Physically we can understand the elevation effect as a proxy for the effect of low-level atmospheric moisture. We can understand the terrain roughness effect more directly in terms of the fact that smooth terrain allows a more direct transport of the moisture into the super-cell updrafts.

The model that includes the elevation variables has a lower AIC value and the difference in log likelihoods is interpreted as a good improvement.

```{r}
AIC(model0); AIC(model1)
2 * (as.numeric(logLik(model1)) - as.numeric(logLik(model0)))/T.ppp25$n
```

| diff | interpretation |
|------|----------------|
| 1    | huge           |
| .1   | large          |
| .01  | good           |
| .001 | okay           |

So we conclude that `model1` is a good improvement over `model0` which is an improvement over a CSR.

Diagnostics.

```{r}
plot(envelope(model1, 
              fun = Kest, 
              nsim = 39, 
              global = TRUE), 
     legend = FALSE)
```

The model is shown in red with the 95% uncertainty bands.

Finally we see if adding a cluster term to the model will improve things further.

```{r}
model2 <- kppm(T.ppp25, 
               trend = ~ Zc + El + Tr, 
               covariates = list(Zc = Zc, El = El, Tr = Tr),
               clusters = "Thomas")
summary(model2)
```

We see that the covariates remain statistically significant although the effect sizes are smaller.

We see with the K function that the model fits the data even better.

```{r}
plot(envelope(model2, 
              fun = Kest, 
              nsim = 39, 
              global = TRUE), 
     legend = FALSE)
```

Simulate point patterns from the model.

```{r}
X <- simulate.kppm(model2, 
                   nsim = 3)

par(mfrow = c(2, 2))
plot(T.ppp25)
plot(X[[1]])
plot(X[[2]])
plot(X[[3]])
```

## Spatial logistic regression

Spatial logistic regression is a popular model for point pattern data. The study domain is divided into a grid of cells; each cell is assigned the value one if it contains at least one event, and zero otherwise.

Then a logistic regression models the presence probability $p = P(Y = 1)$ as a function of explanatory variables $X$ in the (matrix) form $$
\log \frac{p}{1-p} = \beta X
$$ where the left-hand side is the logit (log of the odds ratio) and the $\beta$ are the coefficients to be determined.

If your data are stored as `ppp` objects, a spatial logistic model can be fit directly using functions from the {spatstat} package.

Let's consider a canned example from the package (a good strategy in general).

Consider the locations of 57 copper ore deposits (events) and 146 line segments representing geological 'lineaments.' Lineaments are linear features that consist of geological faults.

It is of interest to be able to predict the probability of a copper ore from the lineament pattern. The data are stored as a list in `copper`. The list contains a `ppp` object for the ore deposits and a `psp` object for the lineaments.

```{r}
library(spatstat)
data(copper)
plot(copper$SouthPoints)
plot(copper$SouthLines, add = TRUE)
```

For convenience we first rotate the events (points and lines) by 90 degrees in the anticlockwise direction and save them as separate objects.

```{r}
C <- spatstat::rotate(copper$SouthPoints, pi/2)
L <- spatstat::rotate(copper$SouthLines, pi/2)
plot(C)
plot(L, add = TRUE)
```

We summarize the planar point pattern `C` that we want to model.

```{r}
summary(C)
```

There are 57 ore deposits over a region of size 5584 square km resulting in an intensity of about .01 ore deposits per square km.

Next we create a distance map of the lineaments to be used as a covariate.

```{r}
D <- distmap(L)
plot(D)
```

Models are fit with the `slrm()` function from the {spatstat} package.

```{r}
model <- slrm(C ~ D)
model
```

The model says that the odds of a copper ore along a lineament (D = 0) is exp(-4.723) = .00888. This is slightly less than the overall intensity of .01.

The model also says that for every one unit (one kilometer) increase in distance from a lineament the expected change in the log odds is .0781 [exp(.0781) = 1.0812] or an 8.1% increase in the odds. Ore deposits are more likely between the lineaments.

The fitted method produces an image (raster) of the window giving the local probability of an ore deposit. The values are the probability of a random ore deposit in each pixel.

```{r}
plot(fitted(model))
plot(C, add = TRUE)
```

Integrating the predictions over the area equals the observed number of ore deposits.

```{r}
sum(fitted(model))
```
