---
title: "Assignment 1"
author: "Your Name"
date: January 15, 2021
output: 
  html_document:
    keep_md: true
editor_options: 
  chunk_output_type: console
---

Due: January 15, 2021, 3pm

Total Points: 85

**1** Assign to the vector object `rn` 100 random values from the standard normal distribution using the code chunk below then answer the questions using {base} R code following the model in 'a'.
```{r}
rn <- rnorm(n = 100)
```

a. What is the value in the 25th element (5)
```{r}
rn[25]
```

b. What are the highest and lowest values (10)
c. What is the average over these 100 values? (5)
d. What is the median over these 100 values? (5)
e. How many values are less than 0? (10)
f. How many values are between 1 and 2? (10)

**2** The Palmer penguins data set is available from https://raw.githubusercontent.com/allisonhorst/palmerpenguins/master/inst/extdata/penguins.csv as a csv file.

a. Import the data as a tabled data frame with the name `penguins` using the `read_csv()` function. (10)
b. Print the number of rows and columns in the data frame. (5)
c. Print the column names. (5)
d. Compute the average bill length in millimeters (`bill_length_mm`) ignoring the missing values. (10)
e. How many missing values are there in the vector containing the bill lengths (`penguins$bill_length_mm`)? (10)

