---
title: "Assignment 3"
author: "Answers"
date: January 29, 2021
output: 
  html_document:
    keep_md: true
editor_options: 
  chunk_output_type: console
---

Due: January 29, 2021, 3pm

Total Points: 40

**1** Use the file `Tornadoes.csv` and create a time series plot of the annual number of tornadoes in Florida since 1950. Use a point and a line layer (15).
```{r}
library(tidyverse)
library(ggplot2)

download.file(url = "http://www.spc.noaa.gov/wcm/data/1950-2019_actual_tornadoes.csv",
              destfile = "Tornadoes.csv")

read_csv(file = "Tornadoes.csv") %>%
  filter(st == "FL") %>%
  group_by(yr) %>%
  summarize(nT = n()) %>%
ggplot(mapping = aes(x = yr, y = nT)) +
  geom_line() +
  geom_point()
```

**2** Create a simple feature data frame from the following information about the Joplin tornado that killed 158 people on May 22, 2011 then determine the length in meters. Hint: use `st_linestring()` and `rbind()` to create the simple feature data frame then use `st_length()` to determine the length. (25)

Start longitude: -94.5932 
Start latitude: 37.0524 
End longitude: -94.2213 
End latitude: 36.9838 
crs: 4326

Attributes:
number of fatalities: 158
name: "Joplin Tornado"

```{r}
# sfg object
Joplin_track <- st_linestring(rbind(c(-94.5932, 37.0524), c(-94.2213, 36.9838)))
# sfc object
Joplin.geometry <- st_sfc(Joplin_track, 
                          crs = 4326)
# data.frame object
Joplin.attributes <- data.frame(name = "Joplin Tornado",
                                nFat = 158)
# sf object
Joplin.sf <- st_sf(Joplin.attributes, 
                   geometry = Joplin.geometry)

Joplin.sf %>%
  st_length()
```