---
title: "Assignment 5"
author: "Answers"
date: February 12, 2021
output: 
  html_document:
    keep_md: true
editor_options: 
  chunk_output_type: console
---

Due: February 12, 2021, 3pm

Total Points: 40

**1** The file `http://myweb.fsu.edu/jelsner/temp/data/police.zip` contains demographic information at the county level across the state of Mississippi.

(a) Import the data as a simple features spatial object and assign it to an object called `PE.sf`. Set the CRS to 4326 using the `set_set_crs()` function. (5)
```{r}
if(!"police" %in% list.files()) {
download.file(url = "http://myweb.fsu.edu/jelsner/temp/data/sids2.zip",
              destfil = "police.zip")
unzip("police.zip")
}

library(sf)

PE.sf <- read_sf(dsn = "police",
                   layer = "police") %>%
  st_set_crs(value = 4326)
```

(b) Using the simple feature data frame `PE.sf` from part (a), create a choropleth map of the percentage of people unemployed in 1980 (variable name `UNEMP`) using functions from the {tmap} package. (5)
```{r}
library(tmap)

tm_shape(PE.sf) +
  tm_polygons(col = "UNEMP")
```

(c) Using the simple feature data frame `PE.sf` from part (a), create a choropleth map of the percentage of people unemployed using functions from the {ggplot2} package. (5)
```{r}
library(ggplot2)

ggplot(data = PE.sf) +
  geom_sf(mapping = aes(fill = UNEMP))
```

**2** With data in the simple feature data frame imported in question (1) above, compute the spatial autocorrelation of the percentage of people unemployed using the `moran()` function from the {spdep} package. Hint: First create a list of neighbors and weights using the `poly2nb()` and `nb2list2()` functions. (25)
```{r}
library(spdep)

nbs <- poly2nb(PE.sf)
wts <- nb2listw(nbs)

m <- length(PE.sf$UNEMP)
s <- Szero(wts)

moran(PE.sf$UNEMP, 
      listw = wts, 
      n = m, 
      S0 = s)
```