---
title: "Assignment 7"
author: "Answers"
date: March 5, 2021
output: 
  html_document:
    keep_md: true
editor_options: 
  chunk_output_type: console
---

Due: March 5, 2021, 3pm

Total Points: 50

The file `1950-2018-torn-initpoint.zip` contains shapefiles of the initial location of all tornado reports in the United States over the period 1950-2018. Use the code below to download the file, unzip it, and import the data as a simple feature data frame called `Torn.sf`.
```{r}
if(!"1950-2018-torn-initpoint" %in% list.files()) {
download.file(url = "http://myweb.fsu.edu/jelsner/temp/data/1950-2018-torn-initpoint.zip",
              destfile = "1950-2018-torn-initpoint.zip")
unzip("1950-2018-torn-initpoint.zip")
}

library(sf)

Torn.sf <- st_read(dsn = "1950-2018-torn-initpoint", 
                   layer = "1950-2018-torn-initpoint")
```

(a) Remove the tornado records (rows) where the EF damage rating is less than 1. Recycle the name `Torn.sf`. Hint: Use the `filter()` function with `mag > 0`. (5)
```{r}
library(tidyverse)

Torn.sf <- Torn.sf %>%
  filter(mag > 0)
```

(b) Transform the CRS of the geometry in `Torn.sf` to an Albers equal-area conic projection centered on the state of Iowa using the ESPG code 26975. (5)
```{r}
Torn.sf <- Torn.sf %>%
  st_transform(crs = 26975)
```

(c) Use functions from the {spatstat} and the {maptools} packages to create an unmarked planar point pattern (`ppp`) object called `T.ppp`. Hints: First use the function `as_Spatial()` to convert the simple feature data frame to an S4 spatial class and then use `as.ppp()` to create the `ppp` object and `unmark()` to remove the marks.  (10)
```{r}
library(spatstat)
library(maptools)

T.ppp <- Torn.sf %>%
  as_Spatial() %>%
  as.ppp() %>%
  unmark()
```

(d) Use functions from the {spatstat} package to create an `owin` object called `W` using the boundary of the state of Iowa. Hints: use the `us_states()` function from the {USAboundaries} package and transform the CRS to that in `Torn.sf` then use `as_Spatial()` and `as.owin()`. (10)
```{r}
library(USAboundaries)

W <- us_states(states = "Iowa") %>%
  st_transform(crs = st_crs(Torn.sf)) %>%
  as_Spatial() %>%
  as.owin()
```

(e) Subset the tornado event locations `ppp` object `T.ppp` by the `owin` object and convert the distance unit to kilometers with the `rescale()` function. (10)
```{r}
T.ppp <- T.ppp[W] %>%
  spatstat::rescale(s = 1000, 
                    unitname = "km")
```

(f) Determine the average intensity of tornado reports for the state as a whole. Then make a map showing the local intensity of the tornado reports using the `plot()` method. Which part of the state has experienced the fewest tornadoes? (10)
```{r}
intensity(T.ppp)
T.ppp %>%
  density() %>%
  plot()

plot(T.ppp, 
     pch = '.', 
     add = TRUE)
```
