---
title: "Assignment 11"
author: "Your Name"
date: April 9, 2021
output: 
  html_document:
    keep_md: true
editor_options: 
  chunk_output_type: console
---

Due: April 16, 2021, 3pm

Total Points: 40

Consider the geodata object `parana` from the {geoR} package containing rainfall values from the state of Paran\`a. The data are average rainfall (mm) over the period May-June.

(a) First plot the observed values using the `plot()` method. Then plot the residuals after a 2nd-order trend is removed from the observed values. (5)
```{r}
library(geoR)

plot(parana)

plot(parana,
     trend = "2nd")
```

(b) Compute and plot an empirical variogram on the residuals after removing the 2nd-order trend. Hint: Use the `variog()` function. (5)
```{r}
plot(variog(parana,
            trend = "2nd"),
     max.dist = 400)
```

(c) Fit a variogram model using the exponential function to the empirical variogram. Assume isotropy. Set the initial nugget to 500, the partial sill to 300 and the range to 100. Hint: Use the `likfit()` function. (5)
```{r}
parana.vm <- likfit(parana, 
                    ini = c(300, 100),
                    nugget = 500,
                    cov.model = "exp", 
                    trend = "2nd")
```

(d) Plot the empirical variogram and add the variogram model curve on the same graph. Hint: To add the variogram model curve use the `lines.variomodel()` function with parameters from the model. (10)
```{r}
plot(variog(parana, 
            trend = "2nd", 
            uvec = seq(0, 300, l = 13)))

lines.variomodel(cov.model = "exp", 
                 cov.pars = c(372, 78), 
                 nug = 381, 
                 max.dist = 300, 
                 col = "red")
```

(e) Create a kriged surface by interpolating the rainfall values to locations on a regular grid. Use the `pred_grid(parana$borders, by = 2)` function to generate the grid. (10)
```{r}
pgrid.df <- pred_grid(parana$borders, by = 2)

ki <- krige.conv(parana, 
                 loc = pgrid.df, 
                 krige = krige.control(trend.d = "2nd", 
                                       trend.l = "2nd",
                                       obj.m = parana.vm))
```

(f) Plot the interpolated surface using the `image()` function with the argument `col = rev(terrain.colors(21))`. (5)
```{r}
image(ki, 
      col = rev(terrain.colors(21)))
```